---
title: "BCR + GEX Integration Tutorial -- Part 2: Seurat Analysis and Annotation"
subtitle: "From Cell Ranger GEX output to an annotated B cell Seurat object"
author: "Nachi Nathan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = TRUE,
  warning = FALSE
)
```

***

# Prerequisites
Before running this part, the following must be in place.

**Input files required**

| File | Where it comes from |
|------|-------------------|
| `P1_LN_bcr_data.tsv` | Part 1 output, or download from Zenodo: https://zenodo.org/records/20287324 |
| `P1_PT_bcr_data.tsv` | Part 1 output, or download from Zenodo: https://zenodo.org/records/20287324 |
| `P1_LN_sample_filtered_feature_bc_matrix.h5` | Cell Ranger GEX output, at `per_sample_outs/<sample>/count/` |
| `P1_PT_sample_filtered_feature_bc_matrix.h5` | Cell Ranger GEX output, at `per_sample_outs/<sample>/count/` |

**R packages required**

```r
install.packages(c("dplyr", "ggplot2", "data.table", "tibble",
                   "viridis", "scales", "patchwork", "pheatmap"))
# Seurat v5
install.packages("Seurat")
# Optional but recommended for faster DEG computation
install.packages("devtools")
devtools::install_github("immunogenomics/presto")
```

**R version:** This tutorial was developed under R 4.3 with Seurat v5. Seurat v4
will not work with the `IntegrateLayers()` workflow used in Step 6.


***

# Overview

This part covers the complete single-cell GEX analysis pipeline, from loading
raw count matrices through to a fully annotated Seurat object. It runs in
sequence -- each step feeds directly into the next -- and produces a single
output file at the end: `integrated_S7_annotated.rds`.

The steps are:

1. **GEX QC** -- load `.h5` count matrices, inspect quality metrics, filter
   low-quality cells.
2. **Doublet removal** -- keep only CD79A-positive cells and remove B cell /
   non-B cell doublets.
3. **Add BCR metadata** -- join the BCR TSV from Part 1 to each Seurat object
   by barcode matching.
4. **Solo GEX analysis** -- per-sample normalization, HVG selection (with IG/TCR
   gene exclusion), PCA, clustering, UMAP, BCR overlay.
5. **Integration** -- RPCA batch correction across samples, joint UMAP,
   joint clustering.
6. **Cluster annotation** -- DotPlot, BCR metadata, DEG analysis, final label
   assignment.

***

# Setup

Set these variables before running anything else. All paths downstream are
derived from these.

```{r setup-session}
suppressPackageStartupMessages({
  library(Seurat)
  library(Matrix)
  library(dplyr)
  library(data.table)
  library(tibble)
  library(ggplot2)
  library(viridis)
  library(scales)
  library(patchwork)
  library(pheatmap)
})

# Root directory -- each sample has its own subfolder here
ROOT_DIR <- "C:/Users/natha/OneDrive/Documents/Immcantation/BCR_Tutorial/Github"

# RDS output directory -- all Seurat objects are written here
RDS_DIR <- file.path(ROOT_DIR, "RDS_Objects")
if (!dir.exists(RDS_DIR)) dir.create(RDS_DIR, recursive = TRUE)

# Sample definitions
samples <- list(
  P1_LN = list(
    sample_id  = "P1_LN",
    h5         = file.path(ROOT_DIR, "P1_LN/P1_LN_sample_filtered_feature_bc_matrix.h5"),
    bcr_tsv    = file.path(ROOT_DIR, "P1_LN/Output/P1_LN_bcr_data.tsv"),
    output_dir = RDS_DIR
  ),
  P1_PT = list(
    sample_id  = "P1_PT",
    h5         = file.path(ROOT_DIR, "P1_PT/P1_PT_sample_filtered_feature_bc_matrix.h5"),
    bcr_tsv    = file.path(ROOT_DIR, "P1_PT/Output/P1_PT_bcr_data.tsv"),
    output_dir = RDS_DIR
  )
)

message("Ready. Samples: ", paste(names(samples), collapse = ", "))
```

***

# Step 2: GEX QC and Filtering

## 2.1 QC thresholds

The thresholds below are applied uniformly to all samples. Inspect the violin
plots in section 2.3 carefully before accepting them for your own data -- the
right values depend on the tissue and library quality.

| Parameter | Value | What it removes |
|-----------|-------|----------------|
| `min_features` | 250 | Empty droplets with almost no detected genes |
| `max_features` | 6,500 | Likely doublets with abnormally high gene counts |
| `min_UMI` | 500 | Low-quality or empty droplets |
| `max_UMI` | 35,000 | Likely doublets with abnormally high UMI counts |
| `filter_mt` | 10% | Damaged or dying cells |

When a cell is damaged or dying, cytoplasmic mRNA leaks out while mitochondrial
transcripts remain protected inside the organelle, producing a characteristic
signature of high mt% alongside low gene and UMI counts. A threshold of 10% is
appropriate for lymphoid tissue. For tumor samples with many metabolically active
cells, a slightly higher threshold may be warranted -- always inspect the
distribution first.

```{r thresholds}
min_features   <- 250
max_features   <- 6500
min_UMI        <- 500
max_UMI        <- 35000
filter_mt      <- 10

mito_pattern   <- "^MT-"   # "^MT-" for human, "^mt-" for mouse
min_cells_gene <- 5        # genes detected in fewer cells than this are dropped
```

## 2.2 Load count matrices and create Seurat objects

We load the `.h5` count matrix for each sample and create a Seurat object.
Mitochondrial percentage is calculated immediately after loading, before any
filtering, so we have the full distribution to inspect.

We also suffix every barcode with the sample name at this stage. Cell Ranger
barcodes are not unique across samples -- both P1_LN and P1_PT contain cells
called `ACGT...TGCA-1` -- and adding the sample ID prevents collisions when
samples are merged later. This matches the barcode suffixing done on the BCR
side in Part 1.

```{r load-counts}
seurat_list <- list()

for (s in samples) {

  message("Loading: ", s$sample_id)

  if (!file.exists(s$h5)) {
    stop("H5 file not found for ", s$sample_id, ":\n  ", s$h5,
         "\nCheck that the path is correct.")
  }

  counts <- Read10X_h5(s$h5)
  if (is.list(counts)) counts <- counts[["Gene Expression"]]

  obj <- CreateSeuratObject(
    counts       = counts,
    project      = s$sample_id,
    min.cells    = min_cells_gene,
    min.features = 0
  )

  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = mito_pattern)
  obj <- RenameCells(obj, new.names = paste0(colnames(obj), "_", s$sample_id))

  message("  Cells: ", ncol(obj), " | Genes: ", nrow(obj))
  seurat_list[[s$sample_id]] <- obj
}
```

The two samples loaded as expected:

| Sample | Cells | Genes |
|--------|-------|-------|
| P1_LN  | 6,061 | 18,521 |
| P1_PT  | 8,067 | 19,889 |

P1_PT has more cells and a slightly larger detected transcriptome, consistent
with the greater cellular complexity of tumor tissue.

## 2.3 Inspect QC metrics before filtering

Look at the violin plots carefully before proceeding. You are looking for a
unimodal distribution of `nFeature_RNA` with a lower tail of empty droplets and
possibly an upper tail of doublets; `percent.mt` clustered low with a tail of
damaged cells; and a linear relationship in the UMI vs genes scatter, with
cells falling well above the line being candidate doublets.

```{r qc-plots-pre, fig.width=10, fig.height=4}
for (s_id in names(seurat_list)) {

  obj <- seurat_list[[s_id]]

  print(
    VlnPlot(obj,
            features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
            pt.size  = 0.1,
            ncol     = 3) &
      ggplot2::labs(title = paste0(s_id, " -- QC metrics (pre-filter)"))
  )

  print(FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "percent.mt") +
    ggplot2::ggtitle(paste0(s_id, " -- UMI vs mt%")))

  print(FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
    ggplot2::ggtitle(paste0(s_id, " -- UMI vs genes detected")))
}
```

```{r qc-summary-pre}
pre_stats <- lapply(names(seurat_list), function(s_id) {
  obj <- seurat_list[[s_id]]
  data.frame(
    Sample        = s_id,
    Cells         = ncol(obj),
    Median_genes  = median(obj$nFeature_RNA),
    Median_UMI    = median(obj$nCount_RNA),
    Median_mt_pct = round(median(obj$percent.mt), 2),
    Max_mt_pct    = round(max(obj$percent.mt), 2),
    Cells_mt_gt10 = sum(obj$percent.mt > 10),
    Cells_gene_hi = sum(obj$nFeature_RNA > max_features),
    Cells_gene_lo = sum(obj$nFeature_RNA < min_features),
    Cells_UMI_hi  = sum(obj$nCount_RNA > max_UMI),
    Cells_UMI_lo  = sum(obj$nCount_RNA < min_UMI)
  )
})
knitr::kable(do.call(rbind, pre_stats), caption = "QC summary before filtering")
```

The two samples have notably different QC profiles, reflecting their biology
rather than any technical problem.

**P1_LN** has a median of 1,356 genes and 3,288 UMIs per cell -- consistent with
a lymph node where many cells are resting with modest transcriptional output.
The mitochondrial distribution has a median of 4.9% with 479 cells (7.9%) above
the 10% threshold, confirming a genuine tail of damaged cells. Very few cells
hit the upper gene or UMI thresholds, suggesting doublets are rare.

**P1_PT** has higher median genes (1,857) and UMIs (5,086) per cell. The mt%
distribution is actually cleaner (median 1.67%), but the maximum of 91.04% shows
some severely damaged cells are present. P1_PT has considerably more high-UMI
cells (212 above 35,000) and more low-gene cells (233 below 250), both typical
of tumor tissue which tends to be harder to dissociate cleanly.

## 2.4 Apply QC filters

```{r filter}
filter_summary <- list()

for (s_id in names(seurat_list)) {

  obj          <- seurat_list[[s_id]]
  cells_before <- ncol(obj)

  obj <- subset(
    obj,
    subset = nFeature_RNA >= min_features &
             nFeature_RNA <= max_features &
             nCount_RNA   >= min_UMI      &
             nCount_RNA   <= max_UMI      &
             percent.mt   <= filter_mt
  )

  cells_after <- ncol(obj)
  message(sprintf(
    "%s -- Cells kept: %d / %d  (%.1f%% retained, %d removed)",
    s_id, cells_after, cells_before,
    100 * cells_after / cells_before,
    cells_before - cells_after
  ))

  filter_summary[[s_id]] <- data.frame(
    Sample        = s_id,
    Cells_before  = cells_before,
    Cells_after   = cells_after,
    Cells_removed = cells_before - cells_after,
    Pct_retained  = round(100 * cells_after / cells_before, 1)
  )

  seurat_list[[s_id]] <- obj
}

knitr::kable(do.call(rbind, filter_summary), caption = "Filtering summary")
```

Both samples retained over 90% of cells -- P1_LN 92.0% (485 removed) and
P1_PT 90.9% (732 removed). Most removed cells were damaged (high mt%) rather
than doublets. The higher absolute number removed from P1_PT reflects both its
larger starting size and the greater proportion of debris typical of tumor tissue.

## 2.5 Inspect QC metrics after filtering

```{r qc-plots-post, fig.width=10, fig.height=4}
for (s_id in names(seurat_list)) {
  print(
    VlnPlot(seurat_list[[s_id]],
            features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
            pt.size  = 0.1,
            ncol     = 3) &
      ggplot2::labs(title = paste0(s_id, " -- QC metrics (post-filter)"))
  )
}
```

```{r qc-summary-post}
post_stats <- lapply(names(seurat_list), function(s_id) {
  obj <- seurat_list[[s_id]]
  data.frame(
    Sample        = s_id,
    Cells         = ncol(obj),
    Median_genes  = median(obj$nFeature_RNA),
    Median_UMI    = median(obj$nCount_RNA),
    Median_mt_pct = round(median(obj$percent.mt), 2),
    Max_mt_pct    = round(max(obj$percent.mt), 2)
  )
})
knitr::kable(do.call(rbind, post_stats), caption = "QC summary after filtering")
```

**A note on Seurat v5:** You may see a message saying `Default search for "data"
layer in "RNA" assay yielded no results; utilizing "counts" layer instead`. This
is expected -- in Seurat v5 the normalised data layer is only created after
`NormalizeData()`, which runs in Step 5. The warning is harmless.

## 2.6 Save post-QC objects

```{r save-step2}
for (s in samples) {
  out_path <- file.path(s$output_dir, paste0(s$sample_id, "_S2_postQC.rds"))
  saveRDS(seurat_list[[s$sample_id]], file = out_path)
  message("Saved: ", out_path)
}
```

***

# Step 3: Doublet Removal and B Cell Enrichment

## 3.1 Why a dedicated doublet removal step?

Standard GEX QC (Step 2) catches many doublets by removing cells with abnormally
high gene or UMI counts. But those filters work at the level of the whole
transcriptome -- they miss a different kind of doublet: a B cell stuck to a T
cell, monocyte, or NK cell inside the same droplet. These "phenotypic doublets"
look perfectly normal by count statistics but carry the wrong GEX profile and
land in unexpected positions during clustering.

Our approach is simple and interpretable. We keep only CD79A-positive cells
(the canonical pan-B cell marker), then flag any cell that co-expresses a
non-B cell marker. This is conservative -- it removes only cells with a strong,
unambiguous signal for a contaminating cell type -- and every removed cell has
a clear reason. For datasets that are not B-cell-enriched, algorithmic doublet
detection (DoubletFinder, Scrublet) on the full object before this step would
be the standard approach.

## 3.2 Helper function

```{r doublet-helper}
# Marker-based doublet detection
# Returns barcodes of cells that are CD79A+ AND express any non-B cell marker
get_doublet_cells <- function(seurat_obj) {

  non_b_markers <- c("CD3E", "CD3D", "CD4", "CD8A",   # T cells
                     "LYZ",  "CST3",                   # Monocytes / myeloid
                     "KLRD1", "KLRB1", "CD247")        # NK cells

  cd79a <- tryCatch(
    FetchData(seurat_obj, vars = "CD79A", layer = "counts")[[1]],
    error = function(e) return(NULL)
  )
  if (is.null(cd79a)) return(NULL)

  doublet_cells <- lapply(non_b_markers, function(marker) {
    tryCatch({
      marker_expr <- FetchData(seurat_obj, vars = marker, layer = "counts")[[1]]
      rownames(seurat_obj@meta.data)[cd79a > 0 & marker_expr > 0]
    }, error = function(e) return(NULL))
  })

  unique(unlist(doublet_cells))
}
```

The non-B marker panel covers the three main contaminating cell types in a
B-cell-enriched suspension: T cells (`CD3E`, `CD3D`, `CD4`, `CD8A`), monocytes
and myeloid cells (`LYZ`, `CST3`), and NK cells (`KLRD1`, `KLRB1`, `CD247`).
Depending on your tissue context you may need to adjust this panel -- for example,
adding epithelial markers for mucosal tissue.

## 3.3 B cell enrichment and doublet removal

```{r filter-doublets}
summary_rows <- list()

for (s_id in names(seurat_list)) {

  message(paste(rep("-", 60), collapse = ""))
  message("Processing: ", s_id)

  obj     <- seurat_list[[s_id]]
  n_total <- ncol(obj)
  message("  Cells in (post-QC):          ", n_total)

  # Keep only B cells (CD79A > 0)
  b_cells <- tryCatch({
    cd79a <- FetchData(obj, vars = "CD79A", layer = "counts")[[1]]
    rownames(obj@meta.data)[cd79a > 0]
  }, error = function(e) {
    message("  WARNING: CD79A not found -- skipping B cell filter.")
    return(colnames(obj))
  })

  obj      <- subset(obj, cells = b_cells)
  n_bcells <- ncol(obj)
  message("  B cells (CD79A+):            ", n_bcells,
          "  (", round(100 * n_bcells / n_total, 1), "% of input)")

  # Identify and remove doublets
  doublets   <- get_doublet_cells(obj)
  n_doublets <- length(doublets)
  message("  Doublets detected:           ", n_doublets)

  if (n_doublets > 0) obj <- subset(obj, cells = setdiff(colnames(obj), doublets))

  n_final <- ncol(obj)
  message("  Cells out (post-doublet):    ", n_final,
          "  (", round(100 * n_final / n_bcells, 1), "% of B cells retained)")

  seurat_list[[s_id]] <- obj

  summary_rows[[s_id]] <- data.frame(
    Sample           = s_id,
    Cells_in         = n_total,
    B_cells_CD79Apos = n_bcells,
    Pct_B_cells      = round(100 * n_bcells / n_total, 1),
    Doublets_removed = n_doublets,
    Cells_out        = n_final,
    Pct_retained     = round(100 * n_final / n_bcells, 1)
  )
}

knitr::kable(do.call(rbind, summary_rows), caption = "Doublet removal summary")
```

The most striking difference between the two samples is the B cell fraction:
46% for P1_LN versus 85.1% for P1_PT. This is not a quality issue -- P1_PT was
FACS-sorted for B cells (CD19+) before sequencing, so the non-B cell fraction
was already depleted at the library preparation stage. P1_LN was sequenced as a
bulk suspension without prior sorting, containing the full cellular composition
of the lymph node. The CD79A filter brings both objects to the same starting
point for downstream analysis.

Doublet rates are similar and expected: 272 removed from P1_LN (10.6% of CD79A+
cells) and 578 from P1_PT (9.3%). Both samples retain just under 90% of their
B cells after doublet removal.

## 3.4 Save post-doublet objects

```{r save-step3}
for (s in samples) {
  out_path <- file.path(s$output_dir, paste0(s$sample_id, "_S3_postDoublet.rds"))
  saveRDS(seurat_list[[s$sample_id]], file = out_path)
  message("Saved: ", out_path)
}
```

***

# Step 4: Adding BCR Metadata

## 4.1 What gets added and why

The BCR TSV from Part 1 has one row per contig. Before joining to Seurat we
reshape it to one row per cell, pulling the relevant columns from the heavy and
light chain rows separately. The metadata columns added are:

| Column | Source | Description |
|--------|--------|-------------|
| `BCR` | heavy chain | TRUE if cell has a detected heavy chain |
| `c_call` | heavy chain | Isotype (IGHM, IGHG1, IGHA1 etc.) |
| `clone_id` | heavy chain | Clone identifier |
| `clone_count` | heavy chain | Number of cells in this clone |
| `v_call` | heavy chain | V gene assignment |
| `j_call` | heavy chain | J gene assignment |
| `mu_freq_H` | heavy chain | Heavy chain SHM frequency |
| `umi_count_H` | heavy chain | Heavy chain UMI count |
| `kappa_lambda` | light chain | IGK or IGL |
| `mu_freq_L` | light chain | Light chain SHM frequency |
| `umi_count_L` | light chain | Light chain UMI count |

Cells with no matching BCR entry get `BCR = FALSE` and `NA` for all other BCR
columns. These are genuine B cells (CD79A+) that did not produce a detected BCR
sequence, which can happen due to dropout or low VDJ sequencing depth.

## 4.2 Load objects and BCR TSVs, join, and save

```{r add-bcr}
summary_rows <- list()

for (s in samples) {

  message(paste(rep("=", 60), collapse = ""))
  message("Processing: ", s$sample_id)

  # Load Seurat object
  rds_path <- file.path(s$output_dir, paste0(s$sample_id, "_S3_postDoublet.rds"))
  if (!file.exists(rds_path)) stop("Cannot find RDS for ", s$sample_id, " -- run Step 3 first.")
  obj <- readRDS(rds_path)
  message("  B cells loaded (post-doublet): ", ncol(obj))

  # Load BCR TSV
  if (!file.exists(s$bcr_tsv)) stop("Cannot find BCR TSV for ", s$sample_id, " -- run Part 1 first.")
  bcr <- data.table::fread(s$bcr_tsv, sep = "\t", data.table = FALSE, nThread = 1)
  message("  BCR TSV loaded: ", nrow(bcr), " rows")

  # Sanity checks: catch any residual duplicate chains
  heavy_rows <- dplyr::filter(bcr, locus == "IGH")
  dup_heavy  <- heavy_rows %>% dplyr::count(cell_id) %>% dplyr::filter(n > 1)
  if (nrow(dup_heavy) > 0) {
    warning("  ", nrow(dup_heavy), " cells have >1 heavy chain -- keeping highest UMI.")
    heavy_rows <- heavy_rows %>%
      dplyr::group_by(cell_id) %>%
      dplyr::slice_max(umi_count, n = 1, with_ties = FALSE) %>%
      dplyr::ungroup()
  } else {
    message("  Sanity check passed: no duplicate heavy chains")
  }

  light_rows <- dplyr::filter(bcr, locus %in% c("IGK", "IGL"))
  dup_light  <- light_rows %>% dplyr::count(cell_id) %>% dplyr::filter(n > 1)
  if (nrow(dup_light) > 0) {
    warning("  ", nrow(dup_light), " cells have >1 light chain -- keeping highest UMI.")
    light_rows <- light_rows %>%
      dplyr::group_by(cell_id) %>%
      dplyr::slice_max(umi_count, n = 1, with_ties = FALSE) %>%
      dplyr::ungroup()
  } else {
    message("  Sanity check passed: no duplicate light chains")
  }

  # Build per-cell metadata table
  heavy_meta <- heavy_rows %>%
    dplyr::transmute(cell_id, c_call, clone_id, clone_count,
                     v_call, j_call, mu_freq_H = mu_freq, umi_count_H = umi_count)

  light_meta <- light_rows %>%
    dplyr::transmute(cell_id,
                     kappa_lambda = ifelse(locus == "IGK", "kappa", "lambda"),
                     mu_freq_L = mu_freq, umi_count_L = umi_count)

  # Check cell ID overlap and report match rate
  b_cells     <- colnames(obj)
  bcr_cells   <- unique(heavy_meta$cell_id)
  n_matched   <- length(intersect(b_cells, bcr_cells))
  n_bcr_only  <- length(setdiff(bcr_cells, b_cells))
  n_gex_only  <- length(setdiff(b_cells, bcr_cells))
  pct_matched <- round(100 * n_matched / length(b_cells), 1)

  message("  BCR matching:")
  message("    B cells in Seurat:          ", length(b_cells))
  message("    Cells with BCR detected:    ", length(bcr_cells))
  message("    Matched:                    ", n_matched, "  (", pct_matched, "% of B cells)")
  message("    BCR only (filtered in GEX): ", n_bcr_only)
  message("    B cells with no BCR:        ", n_gex_only)

  if (pct_matched < 30) {
    warning("Less than 30% of B cells matched to BCR data for ", s$sample_id,
            ". Check that barcode suffixes are consistent between Part 1 and Part 2.")
  }

  # Join BCR metadata onto Seurat cells
  meta_add <- tibble::tibble(cell_id = b_cells) %>%
    dplyr::mutate(BCR = cell_id %in% bcr_cells) %>%
    dplyr::left_join(heavy_meta, by = "cell_id") %>%
    dplyr::left_join(light_meta, by = "cell_id")

  meta_df <- as.data.frame(meta_add)
  rownames(meta_df) <- meta_df$cell_id
  meta_df$cell_id   <- NULL
  obj <- AddMetaData(obj, metadata = meta_df)
  message("  BCR+ cells: ", sum(obj$BCR, na.rm = TRUE))

  # Save
  out_path <- file.path(s$output_dir, paste0(s$sample_id, "_S4_postBCR.rds"))
  saveRDS(obj, file = out_path)
  message("  Saved: ", out_path)

  summary_rows[[s$sample_id]] <- data.frame(
    Sample      = s$sample_id,
    B_cells     = length(b_cells),
    BCR_cells   = length(bcr_cells),
    Matched     = n_matched,
    Pct_matched = pct_matched,
    BCR_plus    = sum(obj$BCR, na.rm = TRUE)
  )
}

knitr::kable(do.call(rbind, summary_rows), caption = "BCR matching summary")
```

P1_LN has a match rate of 90.9% -- expected for an unsorted bulk suspension
where the GEX and VDJ libraries come from the same cells. P1_PT has a lower
match rate of 59.3%. Differences in library depth between GEX and VDJ runs can
cause this -- some cells are well-represented in the GEX library but have
insufficient VDJ sequencing depth to yield a detectable BCR. This is a known
occurrence in multi-library 10x experiments. If you see a match rate below 30%
in your own data, the most likely cause is a barcode suffix mismatch between
Part 1 and Part 2.

## 4.3 Inspect BCR metadata distribution

```{r inspect-bcr}
for (s in samples) {
  obj     <- readRDS(file.path(s$output_dir, paste0(s$sample_id, "_S4_postBCR.rds")))
  bcr_pos <- subset(obj, subset = BCR == TRUE)

  message("\n--- ", s$sample_id, " ---")
  message("BCR TRUE/FALSE:")
  print(table(obj$BCR))

  message("Isotype (BCR+ cells):")
  print(table(bcr_pos$c_call, useNA = "no"))

  message("Top 10 largest clones:")
  top_clones <- bcr_pos@meta.data %>%
    dplyr::filter(!is.na(clone_id)) %>%
    dplyr::distinct(clone_id, clone_count) %>%
    dplyr::arrange(dplyr::desc(clone_count)) %>%
    head(10)
  print(top_clones)
}
```

The isotype and clone size distributions confirm the BCR metadata joined
correctly and are biologically sensible.

**P1_LN** shows a mixed isotype composition with IgM, IgA1, and IgG1 all
represented. The clone size distribution is flat -- the largest clone has only
6 cells -- the polyclonal signature of an ongoing, diverse immune response.

**P1_PT** is strikingly different. IgG1 dominates (2,150 cells). The top clones
contain between roughly 38 and 151 cells each -- highly oligoclonal. This is
the hallmark of antigen-driven clonal selection and expansion in the tumor
microenvironment.

## 4.4 BCR summary plots

```{r bcr-plots, fig.width=8, fig.height=4}
# BCR+ vs BCR- per sample
bcr_counts <- do.call(rbind, lapply(names(samples), function(s_id) {
  obj <- readRDS(file.path(samples[[s_id]]$output_dir,
                           paste0(s_id, "_S4_postBCR.rds")))
  data.frame(Sample = s_id, Status = ifelse(obj$BCR, "BCR+", "BCR-"),
             stringsAsFactors = FALSE)
}))

print(
  ggplot(bcr_counts, aes(x = Sample, fill = Status)) +
    geom_bar(position = "fill") +
    scale_y_continuous(labels = scales::percent_format()) +
    scale_fill_manual(values = c("BCR+" = "#2E86AB", "BCR-" = "#D9D9D9")) +
    labs(title = "B cells with and without detected BCR",
         x = NULL, y = "Percentage of B cells", fill = NULL) +
    theme_classic(base_size = 13)
)

# Isotype distribution per sample
isotype_order <- c("IGHD", "IGHM", "IGHA1", "IGHA2",
                   "IGHG1", "IGHG2", "IGHG3", "IGHG4", "IGHE")

isotype_counts <- do.call(rbind, lapply(names(samples), function(s_id) {
  obj     <- readRDS(file.path(samples[[s_id]]$output_dir,
                               paste0(s_id, "_S4_postBCR.rds")))
  bcr_pos <- subset(obj, subset = BCR == TRUE)
  df <- as.data.frame(table(bcr_pos$c_call), stringsAsFactors = FALSE)
  colnames(df) <- c("Isotype", "Count")
  df$Pct    <- 100 * df$Count / sum(df$Count)
  df$Sample <- s_id
  df
}))

isotype_counts$Isotype <- factor(isotype_counts$Isotype, levels = isotype_order)

print(
  ggplot(isotype_counts, aes(x = Isotype, y = Pct, fill = Sample)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = c("P1_LN" = "#2E86AB", "P1_PT" = "#E84855")) +
    labs(title = "Isotype distribution per sample (BCR+ cells)",
         x = NULL, y = "% of BCR+ cells", fill = "Sample") +
    theme_classic(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
)
```

The isotype plot makes the LN versus PT contrast immediately visible. P1_LN has
a balanced distribution across isotypes while P1_PT is dominated by IgG1 --
a single panel that captures the core immunological difference between the two
tissue compartments.

***

# Step 5: Solo GEX Analysis

## 5.1 Why we exclude IG and TCR variable genes from HVG selection

Before running the analysis loop, this non-default choice deserves explanation.
The standard Seurat workflow selects the 2,000 most variable genes and uses them
for PCA. In B cells this has a specific problem: V(D)J recombination makes the
V gene transcripts extremely variable across cells, but this variance reflects
clonal diversity -- which antibody gene a cell expresses -- rather than
biological cell state.

If we leave these genes in, PCA and UMAP become partly driven by V gene usage.
Cells sharing a V gene get pulled together regardless of transcriptional identity,
and biologically meaningful states like germinal center, plasma cell, or memory
B cell can get distorted.

The fix: run `FindVariableFeatures()` as normal, then remove any gene matching a
V/D/J pattern from the HVG list before scaling and PCA. The genes are **not
removed from the object** -- they remain available for feature plots, DEG
analysis, and everything else. They are only excluded from the features used to
compute the embedding.

```{r gene-filter-setup}
variable_gene_patterns <- c(
  "^IGHV", "^IGKV", "^IGKJ", "^IGLV", "^IGLJ",
  "^IGHD[0-9]", "^IGHJ",
  "^TRAV", "^TRBV", "^TRAJ", "^TRBJ", "^TRBD"
)
variable_gene_regex <- paste(variable_gene_patterns, collapse = "|")
```

One pattern worth clarifying: `^IGHD[0-9]` catches the IGHD diversity genes
(e.g. IGHD3-10), not the IgD isotype transcript. The isotype transcript is
named simply `IGHD` with no number suffix and is not matched by this regex. It
remains in the HVG list if it is variable, which is correct -- IgD expression
is one of the best markers distinguishing naive from class-switched B cells.

## 5.2 Marker panel and color scheme

Used for DotPlots and FeaturePlots throughout Steps 5, 6, and 7.

```{r marker-setup}
b_cell_markers <- c(
  # B cell identity
  "CD79A", "CD19", "MS4A1",
  # Naive
  "IGHD", "IGHM", "IL4R", "TCL1A",
  # Memory
  "CD27", "CD24", "TNFRSF13B",
  # Germinal center
  "BCL6", "AICDA", "MEF2B",
  # Plasmablast / plasma cell
  "JCHAIN", "MZB1", "XBP1",
  # Proliferating
  "MKI67", "UBE2C", "PCNA",
  # Early activation
  "CD83", "MYC", "CCND2", "TRAF4", "MIR155HG",
  # Interferon-stimulated
  "IFI6", "IFI44", "IFIT1", "IFIT2", "IFIT3",
  # Atypical memory / ABCs
  "FCRL4", "FCRL5", "ZBTB32", "ITGAX", "ZEB2",
  # Exhaustion / regulatory
  "HAVCR1", "HAVCR2", "TIGIT", "LAG3", "IL10"
)

isotype_colors <- c(
  "IGHD"  = "#e41a1c", "IGHM"  = "#377eb8", "IGHA1" = "#a65628",
  "IGHA2" = "#f781bf", "IGHG1" = "#4daf4a", "IGHG2" = "#984ea3",
  "IGHG3" = "#ff7f00", "IGHG4" = "#ffff33", "IGHE"  = "#66c2a5"
)
```

## 5.3 Cell cycle gene sets

```{r cc-genes}
s.genes   <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes
```

Cell cycle scores are stored as metadata but not regressed out. In a B cell
dataset, cell cycle state is biologically meaningful -- proliferating cells are
typically dark zone GC B cells undergoing rapid division, and we want this
visible in the UMAP.

## 5.4 Per-sample processing loop

For each sample: normalize, find HVGs, exclude IG/TCR variable genes, scale,
run PCA, select PCs (fewest explaining 80% of variance, floor 10, ceiling 25),
cluster, run UMAP, score cell cycle, generate a full panel of plots, save.

```{r solo-gex, fig.width=10, fig.height=7}
set.seed(42)  # ensures reproducibility of RunPCA, FindNeighbors, FindClusters, RunUMAP, and CellCycleScoring across both samples
var_expl           <- 0.8
cluster_resolution <- 0.5
seurat_list        <- list()

for (s in samples) {

  message(paste(rep("=", 60), collapse = ""))
  message("Processing: ", s$sample_id)

  rds_in <- file.path(s$output_dir, paste0(s$sample_id, "_S4_postBCR.rds"))
  if (!file.exists(rds_in)) stop("Cannot find input RDS for ", s$sample_id, " -- run Step 4 first.")
  obj <- readRDS(rds_in)
  DefaultAssay(obj) <- "RNA"
  message("  Loaded: ", ncol(obj), " cells | ", nrow(obj), " genes")

  # Normalize
  obj <- NormalizeData(obj, normalization.method = "LogNormalize",
                       scale.factor = 10000, verbose = FALSE)

  # Find HVGs and exclude IG/TCR variable genes
  obj <- FindVariableFeatures(obj, selection.method = "vst",
                              nfeatures = 2000, verbose = FALSE)
  genes_to_exclude <- grep(variable_gene_regex, rownames(obj), value = TRUE)
  n_excluded       <- length(intersect(genes_to_exclude, VariableFeatures(obj)))
  message("  Excluding ", n_excluded, " IG/TCR variable genes from HVG list")
  hvg_filtered <- setdiff(VariableFeatures(obj), genes_to_exclude)
  if (length(hvg_filtered) < 50) {
    warning("  Fewer than 50 HVGs remain after filtering. Using unfiltered list.")
    hvg_filtered <- VariableFeatures(obj)
  }
  VariableFeatures(obj) <- hvg_filtered
  message("  HVGs after filtering: ", length(hvg_filtered))

  # Scale and PCA
  obj <- ScaleData(obj, features = hvg_filtered, verbose = FALSE)
  obj <- RunPCA(obj, features = hvg_filtered, npcs = 25, verbose = FALSE)

  eigs    <- obj@reductions$pca@stdev^2
  cum_var <- cumsum(eigs) / sum(eigs)
  n_pc    <- which(cum_var >= var_expl)[1]
  if (is.na(n_pc)) n_pc <- 25
  n_pc    <- max(n_pc, 10); n_pc <- min(n_pc, 25)
  message("  PCs selected: ", n_pc,
          " (explains ", round(100 * cum_var[n_pc], 1), "% of variance)")

  print(
    ElbowPlot(obj, ndims = 25) +
      ggtitle(paste0(s$sample_id, " -- Elbow plot")) +
      geom_vline(xintercept = n_pc, linetype = "dashed", color = "red") +
      labs(caption = paste0("Red dashed line = PC ", n_pc,
                            " (", round(100 * cum_var[n_pc], 1), "% cumulative variance)"))
  )

  # Cluster and UMAP
  obj <- FindNeighbors(obj, dims = 1:n_pc, verbose = FALSE)
  obj <- FindClusters(obj, resolution = cluster_resolution, verbose = FALSE)
  obj <- RunUMAP(obj, dims = 1:n_pc, verbose = FALSE)
  message("  Clusters found: ", length(unique(obj$seurat_clusters)))

  # Cell cycle scoring
  obj <- CellCycleScoring(obj, s.features = s.genes,
                          g2m.features = g2m.genes, set.ident = FALSE)
  obj$Phase <- factor(obj$Phase, levels = c("G1", "S", "G2M"))

  # Plots
  num_clusters <- table(obj$seurat_clusters)
  print(
    DimPlot(obj, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 1.2) +
      scale_color_hue(labels = paste0(names(num_clusters), " (", num_clusters, ")")) +
      ggtitle(paste0(s$sample_id, " -- Clusters (res = ", cluster_resolution, ")")) +
      theme(legend.text = element_text(size = 9))
  )

  print(
    FeaturePlot(obj, reduction = "umap",
                features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
                order = TRUE, pt.size = 1.2) +
      patchwork::plot_annotation(title = paste0(s$sample_id, " -- QC metrics on UMAP"))
  )

  markers_present <- intersect(b_cell_markers, rownames(obj))
  print(
    DotPlot(obj, features = markers_present, cols = c("gray90", "red"), dot.scale = 6) +
      RotatedAxis() +
      ggtitle(paste0(s$sample_id, " -- B cell marker panel")) +
      theme(axis.text.x = element_text(size = 8))
  )

  markers_for_feature <- intersect(
    c("CD79A", "IGHD", "IGHM", "CD27", "BCL6", "AICDA",
      "MKI67", "PRDM1", "XBP1", "FCRL4", "FCRL5"),
    rownames(obj)
  )
  if (length(markers_for_feature) > 0) {
    print(
      FeaturePlot(obj, reduction = "umap", features = markers_for_feature,
                  order = TRUE, pt.size = 1.2) &
        scale_color_gradientn(colors = viridis::inferno(256))
    )
  }

  Idents(obj) <- "Phase"
  print(
    DimPlot(obj, reduction = "umap", label = FALSE, pt.size = 1.2,
            cols = c("G1" = "#AAAAAA", "S" = "#E69F00", "G2M" = "#CC79A7")) +
      ggtitle(paste0(s$sample_id, " -- Cell cycle phase"))
  )
  Idents(obj) <- "seurat_clusters"

  cc_df <- obj@meta.data %>%
    dplyr::select(Phase, seurat_clusters) %>%
    dplyr::group_by(seurat_clusters, Phase) %>%
    dplyr::summarise(n = dplyr::n(), .groups = "drop_last") %>%
    dplyr::mutate(pct = 100 * n / sum(n)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(Phase = factor(Phase, levels = c("G1", "S", "G2M")))

  print(
    ggplot(cc_df, aes(x = seurat_clusters, y = pct, fill = Phase)) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = c("G1" = "#AAAAAA", "S" = "#E69F00", "G2M" = "#CC79A7")) +
      ylab("% of cells") + xlab("Cluster") +
      theme_classic(base_size = 12) +
      ggtitle(paste0(s$sample_id, " -- Cell cycle composition per cluster"))
  )

  # BCR overlay plots
  if ("BCR" %in% colnames(obj@meta.data)) {

    obj_bcr <- subset(obj, subset = !is.na(c_call) & c_call != "")
    if (ncol(obj_bcr) > 0) {
      obj_bcr$c_call <- factor(obj_bcr$c_call, levels = isotype_order)
      present_iso    <- intersect(isotype_order, levels(droplevels(obj_bcr$c_call)))
      print(
        DimPlot(obj_bcr, group.by = "c_call", label = FALSE, pt.size = 1.2,
                cols = isotype_colors[present_iso], order = rev(present_iso)) +
          ggtitle(paste0(s$sample_id, " -- Isotype (BCR+ cells)"))
      )
    }

    if ("mu_freq_H" %in% colnames(obj@meta.data)) {
      print(
        FeaturePlot(obj, features = "mu_freq_H", reduction = "umap",
                    order = TRUE, pt.size = 1.2) +
          scale_color_gradientn(colors = c("#FFFFCC", "#41B6C4", "#0C2C84"),
                                na.value = "grey90") +
          ggtitle(paste0(s$sample_id, " -- SHM frequency, heavy chain"))
      )
    }

    if ("clone_count" %in% colnames(obj@meta.data)) {
      print(
        FeaturePlot(obj, features = "clone_count", reduction = "umap",
                    order = TRUE, pt.size = 1.2) +
          scale_color_gradientn(colors = c("#FFFFCC", "#FD8D3C", "#800026"),
                                na.value = "grey90") +
          ggtitle(paste0(s$sample_id, " -- Clone size"))
      )
    }

    if ("c_call" %in% colnames(obj@meta.data)) {
      iso_df <- obj@meta.data %>%
        dplyr::filter(!is.na(c_call), c_call != "", !is.na(seurat_clusters)) %>%
        dplyr::mutate(c_call          = factor(c_call, levels = isotype_order),
                      seurat_clusters = as.factor(seurat_clusters))
      present_iso2 <- intersect(isotype_order, levels(droplevels(iso_df$c_call)))
      print(
        ggplot(iso_df, aes(x = seurat_clusters, y = 1, fill = c_call)) +
          geom_col(position = "fill") +
          scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
          scale_fill_manual(values = isotype_colors[present_iso2], breaks = present_iso2) +
          ylab("Percentage of BCR+ cells") + xlab("Cluster") +
          theme_classic(base_size = 12) +
          ggtitle(paste0(s$sample_id, " -- Isotype composition per cluster"))
      )
    }
  }

  Idents(obj) <- "seurat_clusters"
  rds_out <- file.path(s$output_dir, paste0(s$sample_id, "_S5_soloProcessed.rds"))
  saveRDS(obj, file = rds_out)
  message("  Saved: ", rds_out)

  seurat_list[[s$sample_id]] <- obj
}
```

## 5.5 Interpreting the P1_LN results

P1_LN yields 2,292 B cells across 6 clusters. The memory B cell clusters
(0, 1, 3, 4) together comprise roughly 74% of cells and are essentially
indistinguishable on the DotPlot -- they share CD27 expression, mixed isotypes,
and no enrichment for activation, GC, or proliferation markers. These cells are
not cycling, not in a germinal center, and not receiving strong activation
signals. Cluster 2 (20.8%) is naive B cells, cleanly marked by TCL1A, IGHM,
IL4R, and IGHD. Cluster 5 is a small plasma cell population marked by JCHAIN,
MZB1, and XBP1.

**There is no germinal center cluster in P1_LN.** The lymph node is where
germinal center reactions are supposed to occur, yet there is no BCL6/AICDA-
enriched cluster here. This is a key finding of the paper: these tumor-draining
lymph nodes are functionally compromised, with B cells that are resting rather
than actively responding.

The BCR data reinforces this. Mean clone size is 1.2 cells and the largest clone
has only 6 cells -- essentially all singletons. SHM levels are moderate and
distributed without clear spatial structure, consistent with a pre-existing
polyclonal memory pool rather than ongoing affinity maturation.

## 5.6 Interpreting the P1_PT results

P1_PT yields 5,665 B cells across 11 clusters -- considerably more cells and
transcriptional diversity than the LN. Cluster 1 (21.2%) is marked by FCRL4,
FCRL5, and ITGAX -- atypical memory B cells (ABCs), which accumulate in settings
of chronic antigen exposure including tumors. Clusters 2, 6, and 10 form the
plasma cell compartment (JCHAIN, MZB1, XBP1). Clusters 4 and 8 are GC B cells
(BCL6, AICDA, MEF2B). Cluster 5 (8.1%) is a clean proliferating cluster
(MKI67, UBE2C, PCNA). Cluster 3 (11.2%) is interferon-stimulated (IFIT1/2/3,
OASL) -- entirely absent from the LN.

The BCR data for P1_PT is in a different league. Mean clone size is 27 cells
and the largest clone has 151 cells. IgG1 dominates (66.4%) and SHM is elevated
(mean 5.1%), the hallmark of antigen-selected, class-switched, somatically
hypermutated B cells.

The LN contains a polyclonal, resting memory B cell pool with no active GC
response and no clonal expansion. The PT contains an oligoclonal,
transcriptionally diverse infiltrate with active GC reactions, expanded plasma
cell and atypical memory populations, a proliferating cluster, and large
antigen-driven clones. This contrast is one of the central findings of the paper
and is fully visible at this pre-integration stage.

***

# Step 6: Integration

## 6.1 Why integration is necessary

Without integration, technical variation between samples -- sequencing depth,
capture efficiency, ambient RNA -- can drive UMAP separation that has nothing to
do with biology. A naive merge of the two objects would almost certainly produce
sample-specific clusters, making it impossible to distinguish a real biological
difference (LN vs PT) from a technical artifact.

We use RPCA (Reciprocal PCA) integration via Seurat v5's `IntegrateLayers()`.
RPCA projects each dataset into the PCA space of the other, identifies mutual
nearest neighbors as integration anchors, and corrects each sample's embedding
toward a shared space. It is conservative: it corrects technical offsets while
preserving real biological differences. It works best when samples share a
meaningful proportion of cell types -- which is exactly the case here, since
both samples contain memory and naive B cells.

## 6.2 Choosing an integration method

Three methods are available in Seurat v5:

**RPCA** (used here): conservative, fast, best for same-platform same-experiment
data with shared cell types.

**CCA**: more aggressive, can overcorrect similar samples but handles very
different conditions better. To switch: change `method = RPCAIntegration` to
`method = CCAIntegration` and update `new.reduction` accordingly.

**Harmony**: operates on PCA embeddings directly, scales well to many samples
(>10) and large datasets (>50k cells). To use: change `method = RPCAIntegration`
to `method = HarmonyIntegration` and install the `harmony` package.

Integration does not modify the count matrix. The `RNA` assay is always
unchanged. All downstream analysis -- differential expression, feature plots --
should use the `RNA` assay.

## 6.3 Load, merge, and prepare layers

```{r load-and-merge}
seurat_list_s5 <- list()
for (s in samples) {
  rds_in <- file.path(s$output_dir, paste0(s$sample_id, "_S5_soloProcessed.rds"))
  if (!file.exists(rds_in)) stop("Cannot find RDS for ", s$sample_id, " -- run Step 5 first.")
  obj <- readRDS(rds_in)
  obj$sample_id <- s$sample_id
  seurat_list_s5[[s$sample_id]] <- obj
  message("Loaded: ", s$sample_id, " -- ", ncol(obj), " cells | ", nrow(obj), " genes")
}

merged <- merge(x = seurat_list_s5[[1]], y = seurat_list_s5[-1])
message("Merged object: ", ncol(merged), " cells | ", nrow(merged), " genes")
```

```{r split-layers}
DefaultAssay(merged) <- "RNA"
merged[["RNA"]] <- JoinLayers(merged[["RNA"]])
merged[["RNA"]] <- split(merged[["RNA"]], f = merged$sample_id)
message("Layers after split: ", paste(Layers(merged), collapse = ", "))
```

## 6.4 Normalize, find HVGs, scale, PCA (unintegrated)

We re-normalize on the merged object because `merge()` in Seurat v5 combines raw
count layers but does not carry over the normalized data layer. The same IG/TCR
variable gene exclusion from Step 5 is applied here.

```{r normalize-hvg-integrated}
set.seed(42)  # ensures reproducibility of RunPCA on the merged object
merged <- NormalizeData(merged, normalization.method = "LogNormalize",
                        scale.factor = 10000, verbose = FALSE)
merged <- FindVariableFeatures(merged, selection.method = "vst",
                               nfeatures = 2000, verbose = FALSE)

genes_to_exclude_m <- grep(variable_gene_regex, rownames(merged), value = TRUE)
n_excluded_m       <- length(intersect(genes_to_exclude_m, VariableFeatures(merged)))
hvg_filtered_m     <- setdiff(VariableFeatures(merged), genes_to_exclude_m)
message("IG/TCR variable genes excluded: ", n_excluded_m)
message("HVGs after filtering: ", length(hvg_filtered_m))
VariableFeatures(merged) <- hvg_filtered_m

merged <- ScaleData(merged, features = hvg_filtered_m, verbose = FALSE)
merged <- RunPCA(merged, features = hvg_filtered_m, npcs = 25, verbose = FALSE)

eigs_m    <- merged@reductions$pca@stdev^2
cum_var_m <- cumsum(eigs_m) / sum(eigs_m)
n_pc_m    <- which(cum_var_m >= 0.8)[1]
if (is.na(n_pc_m)) n_pc_m <- 25
n_pc_m    <- max(n_pc_m, 10); n_pc_m <- min(n_pc_m, 25)
message("PCs selected: ", n_pc_m,
        " (explains ", round(100 * cum_var_m[n_pc_m], 1), "% of variance)")

print(
  ElbowPlot(merged, ndims = 25) +
    ggtitle("Elbow plot -- merged unintegrated") +
    geom_vline(xintercept = n_pc_m, linetype = "dashed", color = "red") +
    labs(caption = paste0("Red dashed line = PC ", n_pc_m,
                          " (", round(100 * cum_var_m[n_pc_m], 1), "% cumulative variance)"))
)
```

## 6.5 Pre-integration UMAP

This UMAP is run on the unintegrated PCA solely to show the batch effect before
correction. It is stored as `umap.unintegrated` and is not used in any downstream
analysis.

```{r umap-unintegrated}
set.seed(42)  # ensures reproducibility of the pre-integration RunUMAP
merged <- RunUMAP(merged, dims = 1:n_pc_m, reduction = "pca",
                  reduction.name = "umap.unintegrated", verbose = FALSE)

print(
  DimPlot(merged, reduction = "umap.unintegrated", group.by = "sample_id",
          pt.size = 0.8) +
    ggtitle("UMAP -- Unintegrated (colored by sample)") +
    theme(plot.title = element_text(hjust = 0.5))
)
```

If the two samples form largely separate clouds, that confirms a batch component
to correct. If they already overlap almost completely, integration will have
little effect but is still worth running.

## 6.6 RPCA integration

```{r integrate}
set.seed(42)  # ensures reproducibility of IntegrateLayers (RPCA anchor finding is stochastic)
options(future.globals.maxSize = 2000 * 1024^2)

merged <- IntegrateLayers(
  object         = merged,
  method         = RPCAIntegration,
  orig.reduction = "pca",
  new.reduction  = "integrated.rpca",
  verbose        = FALSE
)

merged[["RNA"]] <- JoinLayers(merged[["RNA"]])
message("Integration complete. Layers rejoined.")
```

## 6.7 Cluster and UMAP on integrated embedding

Resolution 0.5 produces 11 clusters that map cleanly onto recognizable B cell
states for this dataset. If your clusters look heterogeneous on the DotPlot,
try 0.7 or 1.0. If many clusters look identical, try 0.3 or 0.4.

```{r cluster-umap-integrated}
set.seed(42)  # ensures reproducibility of FindNeighbors, FindClusters, and RunUMAP on the integrated embedding
merged <- FindNeighbors(merged, dims = 1:n_pc_m,
                        reduction = "integrated.rpca", verbose = FALSE)
merged <- FindClusters(merged, resolution = 0.5, verbose = FALSE)
merged <- RunUMAP(merged, dims = 1:n_pc_m,
                  reduction = "integrated.rpca", verbose = FALSE)
message("Clusters found: ", length(unique(merged$seurat_clusters)))
```

## 6.8 Evaluating integration

```{r umap-post-integration}
print(
  DimPlot(merged, reduction = "umap", group.by = "sample_id", pt.size = 0.8) +
    ggtitle("UMAP -- Integrated (colored by sample)") +
    theme(plot.title = element_text(hjust = 0.5))
)

num_clusters_m <- table(merged$seurat_clusters)
print(
  DimPlot(merged, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.8) +
    scale_color_hue(labels = paste0(names(num_clusters_m), " (", num_clusters_m, ")")) +
    ggtitle("UMAP -- Integrated clusters (res = 0.5)") +
    theme(legend.text = element_text(size = 9))
)
```

After successful RPCA integration, cells from P1_LN and P1_PT should intermingle
within shared clusters rather than forming two separate islands. Genuine
biological differences between LN and PT will remain visible as differences in
cluster proportions, not wholesale spatial separation.

## 6.9 Cluster composition by sample

```{r cluster-composition}
comp_df <- merged@meta.data %>%
  dplyr::select(seurat_clusters, sample_id) %>%
  dplyr::group_by(seurat_clusters, sample_id) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop_last") %>%
  dplyr::mutate(pct = 100 * n / sum(n)) %>%
  dplyr::ungroup()

print(
  ggplot(comp_df, aes(x = seurat_clusters, y = pct, fill = sample_id)) +
    geom_bar(stat = "identity") +
    ylab("% of cells") + xlab("Cluster") +
    theme_classic(base_size = 12) +
    ggtitle("Cluster composition by sample")
)
```

## 6.10 Save integrated object

```{r save-integrated}
rds_out_s6 <- file.path(RDS_DIR, "integrated_S6.rds")
saveRDS(merged, file = rds_out_s6)
message("Saved: ", rds_out_s6)
message("Cells: ", ncol(merged), " | Clusters: ", length(unique(merged$seurat_clusters)))
```

***

# Step 7: Cluster Annotation

## 7.1 Load the integrated object

```{r load-step7}
rds_in_s7 <- file.path(RDS_DIR, "integrated_S6.rds")
if (!file.exists(rds_in_s7)) stop("Cannot find integrated RDS -- run Step 6 first.")

obj <- readRDS(rds_in_s7)
DefaultAssay(obj) <- "RNA"
Idents(obj) <- "seurat_clusters"

num_clusters_s7 <- table(obj$seurat_clusters)
message("Loaded: ", ncol(obj), " cells | ", length(num_clusters_s7), " clusters")
```

## 7.2 Cell cycle scoring

Cell cycle phase was computed in Step 5 on the per-sample objects. We check
whether it carried through the merge; if not, we score it here.

```{r cell-cycle-step7}
set.seed(42)  # ensures reproducibility of CellCycleScoring if it runs here
if (!"Phase" %in% colnames(obj@meta.data)) {
  message("Phase column not found -- scoring cell cycle on integrated object...")
  obj <- CellCycleScoring(obj, s.features = s.genes,
                          g2m.features = g2m.genes, set.ident = FALSE)
  message("Cell cycle scoring complete.")
} else {
  message("Phase column already present -- skipping scoring.")
}
obj$Phase <- factor(obj$Phase, levels = c("G1", "S", "G2M"))
Idents(obj) <- "seurat_clusters"
```

## 7.3 UMAP by cluster and by sample

```{r umap-cluster-sample-step7}
print(
  DimPlot(obj, reduction = "umap", label = TRUE, label.size = 5, repel = TRUE) +
    scale_color_hue(labels = paste0(names(num_clusters_s7), "  (n=", num_clusters_s7, ")")) +
    ggtitle("Integrated UMAP -- Clusters") +
    theme(plot.title  = element_text(hjust = 0.5),
          legend.text = element_text(size = 9))
)

print(
  DimPlot(obj, reduction = "umap", group.by = "sample_id",
          label = FALSE, pt.size = 0.8) +
    ggtitle("Integrated UMAP -- Sample") +
    theme(plot.title = element_text(hjust = 0.5))
)
```


## 7.4 B cell marker DotPlot

The DotPlot is the primary annotation tool. Read it row by row and look for the marker group that lights up. Pay attention to both dot size (percent expressing) and color (average expression level). This is the first pass. Several clusters will be immediately classifiable, while others will need additional evidence from BCR metadata and DEG analysis.

Clusters 0, 1, and 6 express memory B cell markers (CD27, CD24, TNFSF13B). Clusters 0 and 6 share this profile with no other distinguishing markers between them and will need DEG analysis to separate. Cluster 1 stands out with FCRL4, FCRL5, and a clear interferon signature (IFIT1, IFIT3, IFI44, MX1), consistent with atypical memory B cells. IFN and stress signaling are known drivers of the atypical MBC program in the literature, which is why these features co-occur in the same population.

Cluster 2 is unambiguous naive: TCL1A, IGHD, IGHM, and IL4R.

Clusters 4 and 5 both express the GC program (BCL6, AICDA, MEF2B). Cluster 5 adds a cycling signature (MKI67, UBE2C, PCNA), so cluster 4 is non-cycling GC and cluster 5 is cycling GC.

Clusters 3, 7, and 8 express plasma cell markers (JCHAIN, MZB1, XBP1). Clusters 3 and 7 are typical PC with no other distinguishing features. Cluster 8 adds cycling markers on top of the PC program, suggesting a plasmablast-like or cycling PC population.

Cluster 9 is a small cluster (32 cells) with no obvious signature. At this size it may not represent a distinct population, and could be the leading edge of a cluster that would resolve with more cells.

```{r dotplot-step7}
markers_present_s7 <- intersect(b_cell_markers, rownames(obj))
print(
  DotPlot(obj, features = markers_present_s7,
          cols = c("gray90", "red"), dot.scale = 6) +
    RotatedAxis() +
    ggtitle("B cell marker panel -- integrated clusters") +
    theme(axis.text.x = element_text(size = 8))
)
```

## 7.5 Sample composition per cluster

```{r sample-composition-step7}
sample_df_s7 <- obj@meta.data %>%
  dplyr::select(sample_id, seurat_clusters) %>%
  dplyr::group_by(seurat_clusters, sample_id) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop_last") %>%
  dplyr::mutate(pct = 100 * n / sum(n)) %>%
  dplyr::ungroup()

print(
  ggplot(sample_df_s7, aes(x = seurat_clusters, y = pct, fill = sample_id)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(labels = scales::label_percent(scale = 1, accuracy = 1)) +
    ylab("% of cluster") + xlab("Cluster") +
    theme_classic(base_size = 12) +
    ggtitle("Sample composition per cluster")
)
```

Clusters 0 and 6 are substantially mixed between LN and PT, consistent with
quiescent memory B cell populations found in both tissues. Cluster 2 is
predominantly LN-derived (~75%), as expected for naive B cells. All remaining
clusters are predominantly or exclusively PT-derived.

## 7.6 FeaturePlots: key marker genes on UMAP

```{r featureplots-step7}
markers_for_feature_s7 <- intersect(
  c("CD79A", "IGHD", "IGHM", "CD27", "BCL6", "AICDA",
    "MKI67", "PRDM1", "XBP1", "FCRL4", "FCRL5"),
  rownames(obj)
)
if (length(markers_for_feature_s7) > 0) {
  print(
    FeaturePlot(obj, reduction = "umap", features = markers_for_feature_s7,
                order = TRUE, pt.size = 0.8) &
      scale_color_gradientn(colors = viridis::inferno(256))
  )
}
```

## 7.7 Isotype composition per cluster

Isotype composition adds a BCR-level layer of evidence that transcriptomics
alone cannot provide. Two clusters that both express CD27 can be distinguished
if one is IgM-dominant (unswitched) and the other is IgG-dominant (class-switched).

Cluster 2 is predominantly IgM/IgD, consistent with its naive identity.
Clusters 0 and 6 show mixed isotype composition. The remaining clusters are
predominantly IgG1-dominant and are also largely PT-derived, highlighting a
tissue-level difference: the LN is enriched for IgM and mixed isotype
populations, while the PT is dominated by class-switched IgG.

```{r isotype-step7}
if ("c_call" %in% colnames(obj@meta.data)) {

  obj_bcr_s7 <- subset(obj, subset = !is.na(c_call) & c_call != "")
  if (ncol(obj_bcr_s7) > 0) {
    obj_bcr_s7$c_call <- factor(obj_bcr_s7$c_call, levels = isotype_order)
    present_iso_s7    <- intersect(isotype_order, levels(droplevels(obj_bcr_s7$c_call)))
    print(
      DimPlot(obj_bcr_s7, group.by = "c_call", label = FALSE, pt.size = 0.8,
              cols = isotype_colors[present_iso_s7], order = rev(present_iso_s7),
              reduction = "umap") +
        ggtitle("Integrated UMAP -- Isotype (BCR+ cells)")
    )
  }

  iso_df_s7 <- obj@meta.data %>%
    dplyr::filter(!is.na(c_call), c_call != "", !is.na(seurat_clusters)) %>%
    dplyr::mutate(c_call          = factor(c_call, levels = isotype_order),
                  seurat_clusters = as.factor(seurat_clusters))
  present_iso_s7b <- intersect(isotype_order, levels(droplevels(iso_df_s7$c_call)))
  print(
    ggplot(iso_df_s7, aes(x = seurat_clusters, y = 1, fill = c_call)) +
      geom_col(position = "fill") +
      scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
      scale_fill_manual(values = isotype_colors[present_iso_s7b],
                        breaks = present_iso_s7b) +
      ylab("Percentage of BCR+ cells") + xlab("Cluster") +
      theme_classic(base_size = 12) +
      ggtitle("Isotype composition per cluster")
  )
}
```

## 7.8 SHM frequency and clonality per cluster

SHM frequency adds a further layer of evidence. Cluster 2 is clearly the lowest
in the dataset with a near-zero median, consistent with its naive identity.
Cluster 1 is notably lower than the other memory clusters (median ~0.03 vs
~0.06 for clusters 0 and 6). The remaining clusters (0, 3, 4, 5, 6, 7, 8) are
broadly similar in the ~0.05-0.06 range. Cluster 9 appears elevated but has
only 32 cells, making the estimate unreliable.

```{r shm-step7}
if ("mu_freq_H" %in% colnames(obj@meta.data)) {
  print(
    FeaturePlot(obj, features = "mu_freq_H", reduction = "umap",
                order = TRUE, pt.size = 0.8) +
      scale_color_gradientn(colors = c("#FFFFCC", "#41B6C4", "#0C2C84"),
                            na.value = "grey90") +
      ggtitle("Integrated UMAP -- SHM frequency, heavy chain")
  )

  shm_df_s7 <- obj@meta.data %>%
    dplyr::filter(!is.na(mu_freq_H)) %>%
    dplyr::mutate(seurat_clusters = as.factor(seurat_clusters))
  print(
    ggplot(shm_df_s7, aes(x = seurat_clusters, y = mu_freq_H)) +
      geom_boxplot(outlier.size = 0.5, fill = "steelblue", alpha = 0.6) +
      ylab("SHM frequency (heavy chain)") + xlab("Cluster") +
      theme_classic(base_size = 12) +
      ggtitle("SHM frequency per cluster")
  )

  print(
    FeaturePlot(obj, features = "clone_count", reduction = "umap",
                order = TRUE, pt.size = 0.8) +
      scale_color_gradientn(colors = c("#FFFFCC", "#FD8D3C", "#800026"),
                            na.value = "grey90") +
      ggtitle("Integrated UMAP -- Clone size")
  )
}
```

## 7.9 Clonality per cluster

Cluster 3 stands out as the most clonal in the dataset with a mean clone size
of ~3.6 and ~85% of cells in expanded clones. Clusters 4, 5, 7, and 8 show
moderate-to-high clonality (mean clone size 2.3-2.8, 74-79% expanded).
Cluster 1 also shows substantial clonality (mean ~2.0, 61% expanded).
Clusters 0, 2, and 6 show near-zero clonality (mean ~1.0, <15% expanded).
Cluster 9 has only 4 cells with BCR data and is uninformative here.

```{r clonality-step7}
if ("clone_id" %in% colnames(obj@meta.data)) {

  clone_df_s7 <- obj@meta.data %>%
    dplyr::filter(!is.na(clone_id), !is.na(seurat_clusters)) %>%
    dplyr::group_by(seurat_clusters) %>%
    dplyr::summarise(
      n_cells         = dplyr::n(),
      n_clones        = dplyr::n_distinct(clone_id),
      clonality       = 1 - (n_clones / n_cells),
      mean_clone_size = n_cells / n_clones,
      max_clone_size  = max(table(clone_id)),
      pct_expanded    = 100 * mean(table(clone_id)[clone_id] > 1),
      .groups = "drop"
    ) %>%
    dplyr::arrange(as.numeric(as.character(seurat_clusters)))

  print(knitr::kable(clone_df_s7, digits = 2, caption = "Clonality per cluster"))

  print(
    ggplot(clone_df_s7, aes(x = as.factor(seurat_clusters), y = mean_clone_size)) +
      geom_bar(stat = "identity", fill = "darkorange", alpha = 0.8) +
      ylab("Mean clone size") + xlab("Cluster") +
      theme_classic(base_size = 12) +
      ggtitle("Mean clone size per cluster")
  )
}
```

## 7.10 Cell cycle per cluster

Clusters 5 and 8 are the clear proliferating populations: cluster 5 shows
near-100% S and G2M phase, and cluster 8 shows ~100% S and G2M.

```{r cell-cycle-plot-step7}
Idents(obj) <- "Phase"
print(
  DimPlot(obj, reduction = "umap", label = FALSE, pt.size = 0.8,
          cols = c("G1" = "#AAAAAA", "S" = "#E69F00", "G2M" = "#CC79A7")) +
    ggtitle("Integrated UMAP -- Cell cycle phase")
)
Idents(obj) <- "seurat_clusters"

cc_df_s7 <- obj@meta.data %>%
  dplyr::select(Phase, seurat_clusters) %>%
  dplyr::group_by(seurat_clusters, Phase) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop_last") %>%
  dplyr::mutate(pct = 100 * n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Phase = factor(Phase, levels = c("G1", "S", "G2M")))

print(
  ggplot(cc_df_s7, aes(x = seurat_clusters, y = pct, fill = Phase)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("G1" = "#AAAAAA", "S" = "#E69F00", "G2M" = "#CC79A7")) +
    ylab("% of cells") + xlab("Cluster") +
    theme_classic(base_size = 12) +
    ggtitle("Cell cycle composition per cluster")
)
```

## 7.11 DEG heatmap: top 5 markers per cluster

The heatmap runs `FindAllMarkers()` and plots the top 5 upregulated genes per
cluster as a z-scored average expression heatmap. This gives an unbiased initial
view of what distinguishes each cluster and provides hints that the pairwise DEG
analysis in the following sections will confirm or refine.

Several clusters show interpretable signals: cluster 1 lights up on FCRL4, CCR1,
and DUSP4, consistent with an atypical/IFN-driven MBC program. Cluster 2 shows
IGHD and CD200, consistent with naive identity. Cluster 3 is dominated by
immunoglobulin V gene transcripts (IGHV3-48, IGHV6-1, IGHV3-21, IGKV1-5,
IGHV3-23), consistent with plasma cells. Cluster 4 shows SERPINA9 and MEF2B
(GC markers). Cluster 5 is dominated by mitotic genes (KIF18B, KIFC1, CDCA3,
SPC25). Cluster 8 shows S-phase histone genes alongside IGLV3-19. Clusters 0
and 6 show weak, non-specific signals from this view alone and are resolved in
section 7.12. Cluster 7 also shows a relatively weak top-5 signal, and cluster
9 is too small (32 cells) for reliable per-cluster DEG signal.

For faster computation, install the `presto` package -- Seurat will use it
automatically:

```r
install.packages("devtools")
devtools::install_github("immunogenomics/presto")
```

```{r deg-heatmap}
all_markers <- FindAllMarkers(
  obj, only.pos = TRUE, min.pct = 0.25,
  logfc.threshold = 0.25, verbose = FALSE
)

top5 <- all_markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 5) %>%
  dplyr::ungroup()

message("FindAllMarkers complete. Total DEGs: ", nrow(all_markers))

if (nrow(top5) > 0) {
  genes_to_plot_hm <- unique(as.character(top5$gene))
  genes_to_plot_hm <- intersect(genes_to_plot_hm, rownames(obj))

  avg_expr <- AverageExpression(
    obj, features = genes_to_plot_hm, group.by = "seurat_clusters",
    assays = "RNA", layer = "data", verbose = FALSE
  )$RNA

  avg_scaled <- t(scale(t(avg_expr)))
  avg_scaled[!is.finite(avg_scaled)] <- 0
  rownames(avg_scaled) <- rownames(avg_expr)
  colnames(avg_scaled) <- colnames(avg_expr)

  gene_order_hm    <- unique(as.character(top5$gene)[as.character(top5$gene) %in% rownames(avg_scaled)])
  cluster_order_hm <- colnames(avg_scaled)[order(as.numeric(gsub("^g", "", colnames(avg_scaled))))]
  avg_scaled       <- avg_scaled[gene_order_hm, cluster_order_hm, drop = FALSE]

  pheatmap::pheatmap(
    avg_scaled, cluster_rows = FALSE, cluster_cols = FALSE,
    color        = colorRampPalette(c("dodgerblue", "white", "firebrick"))(100),
    fontsize_row = 7, fontsize_col = 9, angle_col = 0,
    main         = "Top 5 DEGs per cluster (z-scored average expression)",
    border_color = NA
  )
}
```

## 7.12 Pairwise DEGs: resolving clusters 0, 1, and 2

Clusters 0 and 6 cannot be distinguished from the DotPlot or heatmap alone --
both show a memory B cell profile. We run each against the other two combined.
Cluster 2 is included as a reference point.

```{r deg-cluster-0-1-6}
deg_0_vs_1_6 <- FindMarkers(obj, ident.1 = 0, ident.2 = c(1, 6),
                             min.pct = 0.25, logfc.threshold = 0.1,
                             only.pos = FALSE, verbose = FALSE)
deg_1_vs_0_6 <- FindMarkers(obj, ident.1 = 1, ident.2 = c(0, 6),
                             min.pct = 0.25, logfc.threshold = 0.1,
                             only.pos = FALSE, verbose = FALSE)
deg_6_vs_0_1 <- FindMarkers(obj, ident.1 = 6, ident.2 = c(0, 1),
                             min.pct = 0.25, logfc.threshold = 0.1,
                             only.pos = FALSE, verbose = FALSE)
for (res in list(list(deg_0_vs_1_6, "Cluster 0 vs 1+6"),
                 list(deg_1_vs_0_6, "Cluster 1 vs 0+6"),
                 list(deg_6_vs_0_1, "Cluster 6 vs 0+1"))) {
  df      <- res[[1]]
  label   <- res[[2]]
  df$gene <- rownames(df)
  df      <- df %>% dplyr::arrange(desc(avg_log2FC))
  message("\nTop 30 higher in ", label, ":")
  print(knitr::kable(
    head(df[, c("gene", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")], 30),
    digits = 4, caption = paste("Top 30 DEGs --", label)
  ))
}
```

**Cluster 0** lacks a strong functional identity. Its top markers are immediate
early genes (DUSP1, JUN, JUNB, JUND, NFKBIA), AP-1 family members, and
ribosomal/translation-related genes. The IEG signature can reflect genuine
recent activation, dissociation-induced stress, or both, and on its own does
not define a stable cell state. Combined with near-zero clonality, mixed
isotype composition, and mixed tissue origin, we label this cluster **MBC_Basal**
as the generic memory pool.

**Cluster 1** has the clearest pairwise signal: DHRS9, FCRL4, USP18, ISG15,
IFIT1, IFIT3, DUSP4, RSAD2, CCR1, IFI6, and ITGAX, all with high fold changes
(2.5-3.8). This is a canonical atypical/IFN-driven MBC program. We call this
cluster **MBC_Atypical**.

**Cluster 6** is defined by a stable mature MBC transcription program:
EBF1, BACH2, FOXP1, NOTCH2, and ZEB2 are broadly expressed (50-78% of cells),
alongside CR1, CR2, and ST6GAL1, all significantly upregulated versus
cluster 0. FCRL5 is also elevated. This profile, combined with the lack of
IFN, GC, PC, cycling, or atypical programs, matches the classical CD21+/CR2+
resting memory B cell state. We label this cluster **MBC_CR2pos**.

## 7.13 Pairwise DEGs: distinguishing the PC subtypes

The GC clusters (4 non-cycling, 5 cycling) and the cycling identity of cluster 5
versus 4 are already locked in from the DotPlot and cell cycle plots. The three
PC clusters (3, 7, and 8) all express the core PC program (JCHAIN, MZB1, XBP1),
but the DotPlot already hinted at differences: cluster 8 layers a cycling
signature on top of the PC markers. This step tests each PC cluster against the
other two to identify the specific genes that distinguish them and to confirm
or refine those provisional identities.

```{r deg-pc}
deg_3_vs_78 <- FindMarkers(obj, ident.1 = 3, ident.2 = c(7, 8),
                            min.pct = 0.20, logfc.threshold = 0.1,
                            only.pos = TRUE, verbose = FALSE)
deg_7_vs_38 <- FindMarkers(obj, ident.1 = 7, ident.2 = c(3, 8),
                            min.pct = 0.20, logfc.threshold = 0.1,
                            only.pos = TRUE, verbose = FALSE)
deg_8_vs_37 <- FindMarkers(obj, ident.1 = 8, ident.2 = c(3, 7),
                            min.pct = 0.20, logfc.threshold = 0.1,
                            only.pos = TRUE, verbose = FALSE)
for (res in list(list(deg_3_vs_78, "Cluster 3 vs 7+8"),
                 list(deg_7_vs_38, "Cluster 7 vs 3+8"),
                 list(deg_8_vs_37, "Cluster 8 vs 3+7"))) {
  df      <- res[[1]]
  label   <- res[[2]]
  df$gene <- rownames(df)
  df      <- df %>% dplyr::arrange(desc(avg_log2FC))
  message("\nTop 30 higher in ", label, ":")
  print(knitr::kable(
    head(df[, c("gene", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")], 30),
    digits = 4, caption = paste("Top 30 DEGs --", label)
  ))
}
```

**Cluster 3** is dominated by immunoglobulin heavy and light chain V gene
transcripts (IGHV3-48, IGLV6-57, IGHV3-21, IGKV3-20, IGKV1-5, IGKV3-11, and
many others) alongside ER and secretory pathway machinery (SSR4, PRDX4, ITM2C,
SPAG4, FCGRT, TCN2). This is the canonical antibody-secreting PC signature.
We label this cluster **PC**.

**Cluster 7** shows a strikingly different signature from the other PC
clusters. The top markers are transcription factors and scaffolding genes
(RUNX1, MED13L, WWOX, ATXN1, FNDC3B, SIPA1L3, BCAS3) and notably RGS13 and
RUNX1, both of which are associated with germinal center light zone B cells
and pre-PC differentiation. Bulk Ig V gene transcription is much lower than
cluster 3 and the ER/secretory machinery is largely absent. Combined with the
PC marker expression seen on the DotPlot, this represents a second PC subset
distinct from cluster 3, possibly with retained GC features. We label this
cluster **PC_2**.

**Cluster 8** is dominated by cycling genes with very high fold changes:
S-phase histones (HIST1H3C, HIST1H4F), mitotic kinases and cyclins (CDK1,
PLK1, AURKB, CCNA2, CCNB2), replication machinery (RRM2, TYMS, PCLAF, ORC6,
CDT1, MKI67, TOP2A), and kinetochore/spindle components (KIFC1, CENPA, CENPF,
BIRC5). Combined with PC marker expression on the DotPlot and ~100% S+G2M
cell cycle composition, this is the actively proliferating antibody-secreting
population: **Plasmablast**.

**Cluster 9** is too small (32 cells) for reliable per-cluster DEG signal and
does not show a clear, distinguishing signature. We label it **Unsure** and
treat it cautiously in downstream analyses.

## 7.14 Optional: additional DEG comparisons

Set `deg_mode` to run further comparisons. Left as NULL to skip by default.

```{r deg-finder}
# ---- USER SETTINGS -----------------------------------------------------------
deg_mode     <- NULL   # "one_vs_rest" or "one_vs_one" -- leave NULL to skip
cluster_A    <- 0
cluster_B    <- 1      # one_vs_one only; can be a vector e.g. c(1, 2)
deg_min_pct  <- 0.25
deg_logfc    <- 0.25
deg_only_pos <- TRUE
deg_top_n    <- 20
deg_save_dir <- RDS_DIR
# ------------------------------------------------------------------------------

if (is.null(deg_mode)) {

  message("deg_mode is NULL -- skipping. Set deg_mode to run a comparison.")

} else {

  if (!dir.exists(deg_save_dir)) dir.create(deg_save_dir, recursive = TRUE)
  Idents(obj) <- "seurat_clusters"

  if (deg_mode == "one_vs_rest") {

    deg_result       <- FindMarkers(obj, ident.1 = cluster_A, min.pct = deg_min_pct,
                                    logfc.threshold = deg_logfc, only.pos = deg_only_pos,
                                    verbose = FALSE)
    deg_result$gene  <- rownames(deg_result)
    deg_result       <- deg_result %>% dplyr::arrange(desc(avg_log2FC))
    print(knitr::kable(
      head(deg_result[, c("gene", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")], deg_top_n),
      digits = 4
    ))
    write.table(deg_result, file = file.path(deg_save_dir,
      paste0("DEGs_cluster", cluster_A, "_vs_rest.tsv")),
      sep = "\t", quote = FALSE, row.names = FALSE)

  } else if (deg_mode == "one_vs_one") {

    deg_result            <- FindMarkers(obj, ident.1 = cluster_A, ident.2 = cluster_B,
                                         min.pct = deg_min_pct, logfc.threshold = deg_logfc,
                                         only.pos = deg_only_pos, verbose = FALSE)
    deg_result$gene       <- rownames(deg_result)
    deg_result$cluster    <- cluster_A
    deg_result$comparison <- paste0("cluster", cluster_A, "_vs_cluster",
                                    paste(cluster_B, collapse = "_"))
    deg_result            <- deg_result %>% dplyr::arrange(desc(avg_log2FC))
    print(knitr::kable(
      head(deg_result[, c("gene", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")], deg_top_n),
      digits = 4
    ))
    write.table(deg_result, file = file.path(deg_save_dir,
      paste0("DEGs_cluster", cluster_A, "_vs_cluster",
             paste(cluster_B, collapse = "_"), ".tsv")),
      sep = "\t", quote = FALSE, row.names = FALSE)

  } else {
    stop("Unknown deg_mode: '", deg_mode, "'. Choose from: one_vs_rest, one_vs_one")
  }
}
```

## 7.15 Annotation decisions

After reviewing all plots and DEGs, the final annotation table is:

| Cluster | n    | Label         | Key evidence |
|---------|------|---------------|--------------|
| 0       | 1942 | MBC_Basal     | Near-zero clonality, mixed isotype, immediate early gene program (DUSP1, JUN, JUNB, JUND, NFKBIA); no specific lineage identity |
| 1       | 934  | MBC_Atypical  | FCRL4, ITGAX, USP18, ISG15, IFIT1, IFIT3, DUSP4, RSAD2, CCR1, IFI6; strong atypical + IFN program |
| 2       | 668  | Naive         | TCL1A, IGHD, IGHM, IL4R, CD200; near-zero clonality and SHM; LN-enriched |
| 3       | 665  | PC            | Heavy Ig V gene transcription (IGHV3-48, IGHV3-21, IGKV3-20, many others); ER/secretory machinery (SSR4, PRDX4, ITM2C, SPAG4, FCGRT) |
| 4       | 276  | GC            | BCL6, MEF2B, SERPINA9; non-proliferating GC B cells |
| 5       | 326  | GC_Cycling    | GC markers (BCL6, MEF2B, SERPINA9) plus mitotic program (KIF18B, KIFC1, CDCA3); ~100% S+G2M |
| 6       | 291  | MBC_CR2pos    | CR1, CR2, NOTCH2, ST6GAL1; mature MBC TF program (EBF1, BACH2, FOXP1, ZEB2); LN-biased |
| 7       | 210  | PC_2          | Second PC subset distinct from cluster 3; RGS13, RUNX1, MED13L, WWOX, ATXN1 with retained GC-leaning features |
| 8       | 130  | Plasmablast   | Pure cycling program (HIST1H3C, UBE2C, CCNA2, CDK1, MKI67, TOP2A, AURKB) on top of PC markers; ~100% S+G2M |
| 9       | 32   | Unsure        | No interpretable signature at 32 cells |

```{r cluster-annotations}
# ---- USER SETTINGS -----------------------------------------------------------
cluster_annotations <- c(
  "0" = "MBC_Basal",
  "1" = "MBC_Atypical",
  "2" = "Naive",
  "3" = "PC",
  "4" = "GC",
  "5" = "GC_Cycling",
  "6" = "MBC_CR2pos",
  "7" = "PC_2",
  "8" = "Plasmablast",
  "9" = "Unsure"
)
# ------------------------------------------------------------------------------
```

## 7.16 Apply annotations and validate

```{r apply-annotations}
clusters_in_obj  <- as.character(sort(unique(obj$seurat_clusters)))
clusters_defined <- names(cluster_annotations)
missing_clusters <- setdiff(clusters_in_obj, clusters_defined)
extra_clusters   <- setdiff(clusters_defined, clusters_in_obj)

if (length(missing_clusters) > 0) {
  warning("The following clusters have no annotation: ",
          paste(missing_clusters, collapse = ", "),
          "\nThese cells will be labeled 'Unlabeled'.")
}
if (length(extra_clusters) > 0) {
  message("Note: annotation includes cluster(s) not in object: ",
          paste(extra_clusters, collapse = ", "), " -- ignored.")
}

full_annotation_map               <- cluster_annotations
full_annotation_map[missing_clusters] <- "Unlabeled"
obj$cell_type <- unname(full_annotation_map[as.character(obj$seurat_clusters)])

annotation_summary <- obj@meta.data %>%
  dplyr::group_by(seurat_clusters, cell_type) %>%
  dplyr::summarise(n_cells = dplyr::n(), .groups = "drop") %>%
  dplyr::arrange(as.numeric(as.character(seurat_clusters)))

message("Annotation summary:")
print(knitr::kable(annotation_summary))
```

## 7.17 Annotated UMAPs

```{r annotated-umaps}
annotation_summary$label <- paste0(
  annotation_summary$cell_type, "\n(C", annotation_summary$seurat_clusters,
  ", n=", annotation_summary$n_cells, ")"
)
label_map           <- setNames(annotation_summary$label,
                                as.character(annotation_summary$seurat_clusters))
obj$cell_type_label <- unname(label_map[as.character(obj$seurat_clusters)])

Idents(obj) <- "cell_type_label"
print(
  DimPlot(obj, reduction = "umap", label = TRUE, label.size = 3.5, repel = TRUE) +
    ggtitle("Integrated UMAP -- Annotated (diagnostic)") +
    theme(plot.title = element_text(hjust = 0.5), legend.text = element_text(size = 8)) +
    NoLegend()
)

Idents(obj) <- "cell_type"
print(
  DimPlot(obj, reduction = "umap", label = TRUE, label.size = 4, repel = TRUE) +
    ggtitle("Integrated UMAP -- Cell types") +
    theme(plot.title = element_text(hjust = 0.5)) +
    guides(color = guide_legend(override.aes = list(size = 4)))
)

Idents(obj) <- "seurat_clusters"
```

## 7.18 Annotated DotPlot in canonical cluster order

With annotations assigned, we produce a final DotPlot using the cell type labels
in the canonical cluster order. This mirrors the numbered DotPlot from section
7.4 but with interpretable names on the y-axis.

```{r annotated-dotplot}
cluster_order <- c(
  "Naive", "MBC_Basal", "MBC_CR2pos", "MBC_Atypical",
  "GC", "GC_Cycling", "PC", "PC_2", "Plasmablast", "Unsure"
)

Idents(obj) <- "cell_type"
obj$cell_type <- factor(obj$cell_type, levels = cluster_order)
Idents(obj)   <- "cell_type"

markers_present_annot <- intersect(b_cell_markers, rownames(obj))
print(
  DotPlot(obj, features = markers_present_annot,
          cols = c("gray90", "red"), dot.scale = 6) +
    RotatedAxis() +
    ggtitle("B cell marker panel -- annotated cell types") +
    theme(axis.text.x = element_text(size = 8))
)

Idents(obj) <- "seurat_clusters"
```

## 7.19 Interpreting the annotations

The annotated UMAP shows the full B cell landscape across both tissues.

**Memory B cell compartment (clusters 0, 1, 6):** MBC_Basal and MBC_CR2pos are
both present in LN and PT, both showing low clonality and mixed isotype
composition. MBC_Basal shows an immediate early gene program (DUSP1, JUN, JUNB,
JUND, NFKBIA) without lineage-specific commitment markers and lacks a clear
functional identity beyond CD27+ memory. MBC_CR2pos is defined by complement
receptors (CR1, CR2), NOTCH2, ST6GAL1, and a mature MBC transcription factor
program (EBF1, BACH2, FOXP1, ZEB2), and is LN-biased. MBC_Atypical co-expresses
atypical memory B cell markers (FCRL4, ITGAX) with a strong interferon-stimulated
gene program (USP18, ISG15, IFIT1, IFIT3, RSAD2, IFI6), consistent with the
known link between IFN/stress signaling and the atypical MBC state.

**GC and PC compartment (clusters 3, 4, 5, 7, 8):** The tumor site harbors an
active germinal center response. GC shows the BCL6/MEF2B/SERPINA9 signature
without proliferation. GC_Cycling carries the same GC program with an added
mitotic signature and ~100% S+G2M phase composition. PC is the dominant
antibody-secreting population, defined by heavy Ig V gene transcription and ER
and secretory machinery (SSR4, PRDX4, ITM2C, SPAG4, FCGRT). PC_2 is a smaller,
distinct PC subset with retained GC-leaning features (RGS13, RUNX1) and a
broader transcription factor / scaffolding program rather than dominant
secretory machinery. Plasmablast layers a pure cycling program (HIST1H3C,
UBE2C, CCNA2, CDK1, MKI67, TOP2A, AURKB) on top of the PC markers, with ~100%
S+G2M composition.

**Naive B cells (cluster 2):** TCL1A, IGHD, IGHM, IL4R, and CD200 with
near-zero clonality and SHM, and an IgM/IgD-dominant isotype profile. LN-enriched,
as expected.

**Cluster 9 (Unsure):** 32 cells with no interpretable B cell identity
signature at this cell count. Likely the leading edge of a population that
would resolve with more cells, rather than a distinct biological state.

A note on resolution: this tutorial uses a single patient with one LN and one
PT sample. Some clusters -- particularly Plasmablast (130 cells), PC_2 (210
cells), and Unsure (32 cells) -- have low cell numbers, which limits the
confidence of their annotation. A larger cohort would be needed to fully
characterize these populations. The annotations here are a guide and a starting
point, not a definitive biological conclusion.

***

# Save

```{r save-annotated}
rds_out_s7 <- file.path(RDS_DIR, "integrated_S7_annotated.rds")
saveRDS(obj, file = rds_out_s7)
message("Saved: ", rds_out_s7)
message("Cell types: ", paste(sort(unique(obj$cell_type)), collapse = ", "))
message("Cells: ", ncol(obj))
```

***

*Part 2 complete. Annotated Seurat object saved to `RDS_Objects/`. Continue to
Part 3 (BCR_GEX_Tutorial_Part3.Rmd) for BCR visualization and shared clone
analysis.*
