---
title: "BCR + GEX Integration Tutorial — Part 1: BCR Sequence Processing"
subtitle: "From Immcantation output to a clean, clone-annotated BCR table"
author: "Nachi Nathan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = TRUE,
  warning = FALSE
)
```

---

# Prerequisites

Before running this part of the tutorial, make sure the following are in place.

**VDJ assignment completed**
This tutorial picks up after V(D)J gene assignment has been run on your Cell
Ranger output using the Immcantation Docker container. If you have not done this
yet, work through `00_Docker_Setup_and_VDJ_Assignment.md` first. That guide
covers installing Docker, pulling the Immcantation image, and running
`AssignGenes.py` and `MakeDb.py` to produce the AIRR TSV files used here.

**Input files required**

| File | Where it comes from |
|------|-------------------|
| `<sample>_airr_db-pass.tsv` | Output of `MakeDb.py` from the Docker step |
| `filtered_contig_annotations.csv` | Cell Ranger VDJ output, at `per_sample_outs/<sample>/vdj_b/` |
| IMGT germline reference (`human/vdj/`) | Downloaded via fetch_imgtdb.sh from the Immcantation repository (see instructions below) |

**Downloading the IMGT germline reference**

Run the following commands in a terminal (not an R session) once before running
the tutorial. This clones the Immcantation repository and downloads the
formatted IMGT germline sequences to a folder of your choice.

```bash
git clone https://github.com/immcantation/immcantation
immcantation/scripts/fetch_imgtdb.sh -o germlines
```

The human VDJ reference will be at `germlines/human/vdj/`. Set `IMGT_REF_DIR`
in the paths chunk to point to this folder.

**R packages required**

```r
install.packages(c("dplyr", "ggplot2", "readr", "stringr"))

# Bioconductor / CRAN immunology packages
install.packages("BiocManager")
BiocManager::install(c("airr", "alakazam", "dowser", "scoper", "shazam"))
```

**R version:** This tutorial was developed under R 4.3. Any R 4.x version should work.

---

# Overview

This part of the tutorial covers BCR sequence processing using the Immcantation
R toolkit. Starting from the AIRR TSV produced by IgBLAST, we QC and filter
the sequences, assign B cell clones, reconstruct germline sequences, and
calculate somatic hypermutation (SHM). The output is a clean, annotated TSV
per sample that feeds into Part 2 (Seurat integration) and Part 3 (repertoire
visualization).

The dataset comes from Nathan et al., *Immunity* (2026), a study of
tumor-draining lymph nodes in ovarian cancer patients. We use two samples from
the same patient: a lymph node (LN) and a primary tumor (PT). This pairing
lets us demonstrate multi-sample handling and cross-tissue clone tracking.

The steps in this part are:

1. **Load and filter** -- load the AIRR TSV and Cell Ranger annotations, append
   sample IDs to barcodes, and apply four sequential QC filters.
2. **Clonal threshold** -- determine the CDR3 distance threshold separating
   clonally related from unrelated sequences.
3. **Clone assignment** -- hierarchical clustering to define clones, followed
   by human-readable clone ID encoding.
4. **Germline reconstruction and SHM** -- reconstruct unmutated germlines and
   calculate per-sequence mutation frequency.
5. **Save** -- write a clean TSV per sample for downstream use.
6. **Repertoire overview** -- rank-abundance, clone size, and SHM plots to
   characterize each sample before moving to GEX.

---

# Setup

## Paths

Set these variables to match your file structure before running anything else.
Every path downstream is derived from these, so this is the only place you need
to make changes.

```{r paths}
# File containing BCR helper functions (BCR_functions.R from the tutorial repo)
BCR_FUNCTIONS_FILE <- "C:/Users/natha/OneDrive/Documents/Immcantation/BCR_Tutorial/Github/BCR_functions.R"

# IMGT germline reference directory (human VDJ)
# This is the folder you copied out of the Docker container during the VDJ step.
# It should contain files like IGHV.fasta, IGHD.fasta, IGHJ.fasta, etc.
IMGT_REF_DIR <- "C:/Users/natha/OneDrive/Documents/Immcantation/BCR_Tutorial/Github/human/vdj"

# Root directory — each sample has its own subfolder here
ROOT_DIR <- "C:/Users/natha/OneDrive/Documents/Immcantation/BCR_Tutorial/Github"

# Sample definitions
# Each entry needs: a sample ID, the path to the AIRR TSV, the path to the
# Cell Ranger filtered_contig_annotations.csv, and an output folder.
samples <- list(
  P1_LN = list(
    sample_id        = "P1_LN",
    airr_file        = file.path(ROOT_DIR, "P1_LN/results/P1_LN_airr_db-pass.tsv"),
    annotations_file = file.path(ROOT_DIR, "P1_LN/P1_LN_filtered_contig_annotations.csv"),
    output_dir       = file.path(ROOT_DIR, "P1_LN/Output")
  ),
  P1_PT = list(
    sample_id        = "P1_PT",
    airr_file        = file.path(ROOT_DIR, "P1_PT/results/P1_PT_airr_db-pass.tsv"),
    annotations_file = file.path(ROOT_DIR, "P1_PT/P1_PT_filtered_contig_annotations.csv"),
    output_dir       = file.path(ROOT_DIR, "P1_PT/Output")
  )
)

# Create output directories if they don't exist
for (s in samples) {
  if (!dir.exists(s$output_dir)) dir.create(s$output_dir, recursive = TRUE)
}
```

## Packages

```{r packages, message=FALSE}
suppressPackageStartupMessages({
  library(airr)
  library(alakazam)
  library(dplyr)
  library(dowser)
  library(ggplot2)
  library(readr)
  library(scoper)
  library(shazam)
  library(stringr)
})
```

## Load functions

```{r load-functions}
source(BCR_FUNCTIONS_FILE)
```

---

# Step 1: BCR QC and Clonal Assignment

## 1.1 Load and filter

We load the AIRR TSV from Immcantation and the `filtered_contig_annotations.csv`
from Cell Ranger. The annotations file provides UMI counts per contig, which we
need later to resolve cells with multiple light chains.

The first thing we do after loading is append the sample ID to every barcode.
Cell Ranger barcodes are not unique across samples -- both samples will have
cells called `ACGT...TGCA-1` -- so without a suffix they collide when samples
are combined downstream. Adding the sample ID at this stage keeps the IDs
unambiguous through every subsequent step.

After loading, we apply four sequential QC filters:

1. Remove non-productive sequences (frameshifts, stop codons)
2. Remove cells with more than one heavy chain -- these are likely doublets
3. Remove light chains with no paired heavy chain
4. For cells with multiple light chains, keep only the one with the highest UMI count

```{r load-and-filter}
bcr_list <- list()

for (s in samples) {
  message("Loading and filtering: ", s$sample_id)

  bcr <- process_bcr_data(
    airr_file        = s$airr_file,
    annotations_file = s$annotations_file,
    sample_id        = s$sample_id
  )
  bcr <- filter_bcr_data(bcr)

  bcr_list[[s$sample_id]] <- bcr
}
```

The QC summary above shows how many sequences were removed at each step and
the final chain pairing breakdown. A proportion of heavy-only cells is expected
-- these are cells where the light chain contig did not pass Cell Ranger's
filters and are kept in the table for clone assignment purposes.

## 1.2 Clonal threshold

To define clones we use hierarchical clustering on the heavy chain CDR3
sequences. The key parameter is the distance threshold -- sequences within this
threshold of each other (measured as normalized Hamming distance on the CDR3)
are assigned to the same clone.

We determine the threshold automatically using SHazaM's Gaussian mixture model
(GMM). The model fits two distributions to the pairwise nearest-neighbor
distance histogram: one for within-clone pairs (short distances) and one for
between-clone pairs (longer distances). The threshold is placed at the point
that best separates the two.

Inspect the plots below for each sample. The two samples look quite different,
and both patterns are biologically meaningful:

- **P1_LN** shows a classic bimodal distribution with a clear valley. The
  threshold lands right in the valley -- this is the ideal case.
- **P1_PT** shows a heavily skewed distribution with almost no second peak.
  This reflects a clonally expanded tumor sample: most B cells belong to a
  small number of large clones, so the majority of pairwise comparisons are
  within-clone (short distances) and there are too few between-clone pairs to
  form a visible second peak. The automatic threshold is still reasonable, but
  the shape of the distribution is itself informative about the biology.

If your data shows a less clear separation and the automatic threshold looks
wrong, you can override it by setting a numeric value in `USER_THRESHOLD` below.

```{r clonal-threshold, fig.width=7, fig.height=4}
# ---- User threshold override ------------------------------------------------
# By default the threshold is determined automatically by the GMM (NULL).
# If the automatic threshold looks wrong for a sample, set a numeric value here
# (e.g. 0.15) to override it. The override will appear as a red dashed line
# on the distance histogram so you can verify it visually.
#
# When to override: if the distribution is unimodal (no clear valley) and the
# auto threshold lands in an obviously wrong place, try a value in the
# 0.10-0.20 range and re-run this chunk.
USER_THRESHOLD <- list(
  P1_LN = NULL,   # NULL = use automatic threshold
  P1_PT = NULL    # NULL = use automatic threshold
)
# -----------------------------------------------------------------------------

threshold_list <- list()

for (s_id in names(bcr_list)) {

  thr_res   <- plot_clonal_thresholds(bcr_list[[s_id]],
                                      user_defined_threshold = USER_THRESHOLD[[s_id]])
  threshold <- if (!is.null(USER_THRESHOLD[[s_id]])) USER_THRESHOLD[[s_id]] else thr_res$auto_threshold
  message(if (!is.null(USER_THRESHOLD[[s_id]])) "User-defined" else "Auto",
          " threshold for ", s_id, ": ", signif(threshold, 4))

  print(thr_res$shazam_plot +
    ggplot2::ggtitle(paste("GMM threshold fit --", s_id)))

  threshold_list[[s_id]] <- threshold
}
```

## 1.3 Clone assignment

With the threshold in hand we run hierarchical clonal clustering. Two settings
are worth noting:

- `only_heavy = TRUE`: clonal grouping is driven by the heavy chain. This is
  standard practice -- the heavy chain CDR3 carries more sequence diversity and
  is more informative for clone definition.
- `split_light = TRUE`: after heavy-chain clustering, any clone containing cells
  with different light chains is split into sub-clones. This resolves cases
  where two cells share the same heavy chain rearrangement but carry different
  receptors.

After assigning clones we replace the default numeric clone IDs with
informative strings. Each clone ID encodes four pieces of information: the
sample ID, a unique random code, the clone size, and the heavy chain isotype
composition. For example, `P1_LN_aB3x_14_G1M` is a clone from the LN sample
with 14 cells carrying both IgG1 and IgM. The random code ensures uniqueness
when data from multiple samples are later combined -- no two clones from
different samples can accidentally share an ID.

```{r clone-assignment}
for (s_id in names(bcr_list)) {

  bcr <- bcr_list[[s_id]]
  bcr <- define_clones_basic(bcr, threshold = threshold_list[[s_id]], sample_id = s_id)
  bcr <- annotate_clone_ids(bcr, sample_id = s_id)
  qc_clone_assignment(bcr, bcr)

  bcr_list[[s_id]] <- bcr
}
```

## 1.4 Germline reconstruction and somatic hypermutation

To measure somatic hypermutation (SHM) we need the unmutated germline sequence
each B cell started from. We reconstruct this using the IMGT reference database
via `dowser::createGermlines()`.

Once germlines are in hand, `shazam::observedMutations()` counts the differences
between each sequence and its germline across the V gene region. We calculate
both mutation frequency (proportion of positions mutated) and raw mutation count,
stored as `mu_freq` and `mu_count` in the output. These will be used for SHM
analysis in Part 3.

This is the slowest step in the pipeline -- expect a minute or two per sample.

```{r germlines-and-mutations}
for (s_id in names(bcr_list)) {
  message("Reconstructing germlines for: ", s_id)

  mut_outputs <- reconstruct_germlines_and_mutations(
    bcr_list[[s_id]],
    reference_dir = IMGT_REF_DIR,
    nproc         = 1
  )

  bcr_list[[s_id]] <- mut_outputs$results_with_mut
}
```

## 1.5 Save

We save a TSV for each sample. These files are the input for Step 4 in Part 2
(Add BCR to Seurat), after GEX QC and doublet removal are complete. They are
also used directly in Part 3 for repertoire visualization.

```{r save-bcr}
for (s in samples) {
  bcr_tsv <- file.path(s$output_dir, paste0(s$sample_id, "_bcr_data.tsv"))
  write.table(bcr_list[[s$sample_id]], file = bcr_tsv,
              sep = "\t", row.names = FALSE, quote = FALSE)
  message("Saved: ", bcr_tsv)
}
```

## 1.6 Repertoire overview

Before moving to GEX it is worth getting a quick overview of each sample's
repertoire. Three plots are shown for each sample:

- **Rank-abundance**: each point is a clone ranked by size. A flat curve means
  many clones of similar size (polyclonal). A steep early drop-off means a few
  dominant clones (oligoclonal).
- **Clone size distribution**: how many clones have 1 cell, 2 cells, 3 cells,
  and so on.
- **Mutation frequency by isotype**: SHM frequency split by heavy chain
  isotype. IgM should be near zero (naive/unswitched); IgG and IgA should be
  higher, reflecting affinity maturation through class switching.

```{r repertoire-plots, fig.width=8, fig.height=4}
for (s_id in names(bcr_list)) {

  bcr <- bcr_list[[s_id]]

  clone_plots <- visualize_clones(bcr)
  print(clone_plots$rank_abundance_plot +
    ggplot2::ggtitle(paste("Rank-abundance --", s_id)))
  print(clone_plots$clone_size_plot +
    ggplot2::ggtitle(paste("Clone size distribution --", s_id)))

  mf_plots <- plot_mutation_frequencies(bcr)
  print(mf_plots$by_isotype +
    ggplot2::ggtitle(paste("Mutation frequency by isotype --", s_id)))
}
```

### What the plots show

The two samples show a striking contrast that reflects their biology:

**P1_LN (lymph node)** is a polyclonal, actively diversifying repertoire. The
rank-abundance curve is flat, the clone size distribution is dominated by
singletons, and the mutation frequency plot shows the expected isotype
hierarchy -- IgM near zero, IgG and IgA progressively higher. This is
consistent with an active germinal center producing a broad, maturing B cell
response.

**P1_PT (primary tumor)** is oligoclonal and more contracted. The top clone
reaches roughly 3.3% abundance, the clone size distribution has a long tail
extending to clones of 100+ cells, and mutation frequencies are lower and more
compressed across isotypes. This is consistent with a selected, expanded
population that has largely exited the germinal center reaction rather than one
that is actively diversifying.

This contrast -- a diverse, functional LN repertoire alongside a contracted,
oligoclonal tumor repertoire -- is one of the central observations of the
accompanying paper. In Parts 2 and 3 we will be able to ask which cell types
the expanded tumor clones belong to.

---

*Part 1 complete. TSV files saved to the Output folder for each sample.
Continue to Part 2 (BCR_GEX_Tutorial_Part2.Rmd) for GEX QC, integration,
and cluster annotation.*
