---
title: "BCR + GEX Integration Tutorial -- Part 3: Repertoire Visualization"
subtitle: "Clonality donuts, phylogenetic trees, and shared clone analysis"
author: "Nachi Nathan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = TRUE,
  warning = FALSE
)
```

---

# Prerequisites
Before running this part, the following must be in place.

**Input files required**

| File | Where it comes from |
|------|-------------|
| `P1_LN_bcr_data.tsv` | Part 1 output, or download from Zenodo: https://zenodo.org/records/20287324 |
| `P1_PT_bcr_data.tsv` | Part 1 output, or download from Zenodo: https://zenodo.org/records/20287324 |
| `integrated_S7_annotated.rds` | Part 2 output, or download from Zenodo: https://zenodo.org/records/20287324 |
| `BCR_viz_functions.R` | This tutorial repo, at `ROOT_DIR/` |

**IQ-TREE 2 (required for phylogenetic trees only)**
The tree section uses the dowser package, which calls IQ-TREE 2 externally to
build the tree topology. IQ-TREE 2 must be installed before running
Section 8.4. The donut plot sections (8.1--8.3) and all of Step 9 do not
require IQ-TREE and can be run independently.

1. Download the IQ-TREE 2 binary for your operating system from:
   https://github.com/Cibiv/IQ-TREE/releases
   (look for the latest release and download the zip for your OS)
2. Unzip and note the path to the executable:
   - Windows: `C:/tools/iqtree2/bin/iqtree2.exe`
   - macOS/Linux: `/usr/local/bin/iqtree2`
3. Either add the bin folder to your system PATH so it can be called as
   `iqtree2`, or supply the full path to the `exec` argument in
   `build_bcr_trees()` as shown in Section 8.4.

**R packages required**

```r
install.packages(c("dplyr", "ggplot2", "data.table", "writexl", "tibble",
                   "tidyr", "patchwork", "scales", "stringr",
                   "dowser", "ape", "RColorBrewer", "ggraph"))
install.packages("Seurat")
# scRepertoire
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("scRepertoire")
# ggtree is on Bioconductor -- cannot be installed with install.packages()
BiocManager::install("ggtree")
```

---

# Overview

This part covers BCR repertoire visualization using two tool ecosystems.
Step 8 uses Immcantation and custom functions to visualize clonal structure
within each sample -- donut plots at the whole-sample, per-isotype, and
per-cluster level, followed by germline-rooted phylogenetic trees for the top
expanded clones. Step 9 uses scRepertoire to ask a different question: are any
clones shared between the LN and PT samples from the same patient?

The dataset comes from Nathan et al., *Immunity* (2026). The two samples used
throughout this tutorial -- P1_LN and P1_PT -- are the lymph node and primary
tumor from the same ovarian cancer patient.

---

# Setup

Set these variables before running anything else. All paths downstream are
derived from these.

```{r setup-session}
suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(ggplot2)
  library(data.table)
  library(writexl)
  library(tibble)
  library(tidyr)
  library(patchwork)
  library(scales)
  library(stringr)
  library(dowser)
  library(ggtree)
  library(ape)
  library(RColorBrewer)
  library(scRepertoire)
  library(ggraph)
})

# Baseline RNG seed for the session. Individual stochastic chunks (clonalNetwork)
# are also seeded immediately before they run, so they remain reproducible when
# executed in isolation.
set.seed(42)

ROOT_DIR <- "C:/Users/natha/OneDrive/Documents/Immcantation/BCR_Tutorial/Github"

# BCR_viz_functions.R lives in ROOT_DIR alongside the tutorial Rmd files
source(file.path(ROOT_DIR, "BCR_viz_functions.R"))

# Input: integrated annotated Seurat object from Part 2
rds_in <- file.path(ROOT_DIR, "RDS_Objects", "integrated_S7_annotated.rds")

# Input: per-sample BCR TSV files from Part 1
bcr_tsv <- list(
  P1_LN = file.path(ROOT_DIR, "P1_LN", "Output", "P1_LN_bcr_data.tsv"),
  P1_PT = file.path(ROOT_DIR, "P1_PT", "Output", "P1_PT_bcr_data.tsv")
)

# Output: annotated BCR files, summary tables, and plots
out_dir  <- file.path(ROOT_DIR, "BCR_Data", "Annotated")
plot_dir <- file.path(ROOT_DIR, "Plots", "Part3")
rds_dir  <- file.path(ROOT_DIR, "RDS_Objects")
dir.create(out_dir,  recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(rds_dir,  recursive = TRUE, showWarnings = FALSE)

message("Ready.")
```

---

# Step 8: BCR Visualization

---

## 8.1 Load the annotated Seurat object and extract cell type labels

We load the integrated object from Part 2 and extract the
barcode-to-cell-type mapping from its metadata. We only need three columns:
the cell barcode, the `cell_type` label, and the `sample_id`. The barcodes in
the integrated object carry the sample suffix (e.g. `ACGT...TGCA-1_P1_LN`)
because this was appended during Part 2. The BCR `cell_id` column carries the
same suffix from Part 1. The join in Step 8.2 therefore works directly without
any barcode manipulation.

Not all BCR cells will match a Seurat barcode. Cells that were filtered out
during GEX QC (low quality, doublets) in Part 2 are present in the BCR data
but absent from the Seurat object. These will have `cell_type = NA` and are
retained in the BCR data for completeness. They appear in combined and
per-isotype donuts -- their BCR sequences are valid -- but are excluded from
per-cluster donuts and from tree tip labels, since they have no cluster
assignment. In this dataset, 87.8% of P1_LN heavy chain cells and 78.1% of
P1_PT heavy chain cells matched a Seurat barcode. Match rates vary depending
on GEX QC stringency and are expected to be somewhat lower for the tumor
sample, where more cells are filtered during QC.

```{r load-seurat}
if (!file.exists(rds_in)) {
  stop("Cannot find integrated RDS at:\n  ", rds_in,
       "\nRun Part 2 first.")
}

obj <- readRDS(rds_in)
message("Integrated object loaded: ", ncol(obj), " cells | ",
        length(unique(obj$cell_type)), " cell types")

# Extract barcode -> cell_type table from Seurat metadata
# rownames of @meta.data are the cell barcodes
cell_type_map <- obj@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  dplyr::select(cell_id, cell_type, sample_id)

message("Cell type distribution:")
print(table(cell_type_map$cell_type))
```

---

## 8.2 Join cell type onto BCR data

For each sample we load the BCR TSV from Part 1 and left-join the `cell_type`
column by `cell_id`. The result is a long-format data frame with one row per
contig (heavy and light chains in separate rows). This long format is required
by all visualization functions and by dowser for tree building.

We use `data.table::fread()` to load the TSV files. These files can be large
and `fread` is substantially faster than `read_tsv` for files stored on
OneDrive, which syncs in the background during reading and can cause `read_tsv`
to hang for several minutes.

```{r join-cell-type}
bcr_annotated <- list()

for (sid in names(bcr_tsv)) {

  tsv_path <- bcr_tsv[[sid]]

  if (!file.exists(tsv_path)) {
    stop("Cannot find BCR TSV for ", sid, " at:\n  ", tsv_path,
         "\nRun Part 1 first.")
  }

  message("\n--- Processing: ", sid, " ---")

  # fread is used here rather than read_tsv for speed on OneDrive-backed paths
  bcr <- data.table::fread(tsv_path, sep = "\t", data.table = FALSE)
  message("  BCR contigs loaded: ", nrow(bcr))

  # Filter cell_type_map to this sample and join by cell_id
  ct_this_sample <- dplyr::filter(cell_type_map, sample_id == sid) %>%
    dplyr::select(cell_id, cell_type)

  bcr <- dplyr::left_join(bcr, ct_this_sample, by = "cell_id")

  # Report match rate on heavy chains only (one per cell)
  n_matched   <- sum(!is.na(bcr$cell_type[bcr$locus == "IGH"]))
  n_igh_total <- sum(bcr$locus == "IGH")
  pct_matched <- round(100 * n_matched / n_igh_total, 1)

  message("  Cell type matched: ", n_matched, " / ", n_igh_total,
          " heavy chain cells (", pct_matched, "%)")

  # Match rates below 50% suggest a barcode format mismatch between
  # Part 1 and Part 2. Check that sample suffixes are consistent.
  if (pct_matched < 50) {
    warning("  Less than 50% of heavy chain cells matched to a cell type. ",
            "Check that barcode suffixes are consistent between Part 1 and Part 2.")
  }

  # Cells not in the Seurat object are left as NA -- they were filtered out
  # during GEX QC and have no valid cluster assignment. They are retained in
  # the BCR data because their BCR sequences are valid, but they will be
  # excluded from per-cluster plots and tree tip labels automatically.

  # Compute VDJ duplicate counts within each clone.
  # For each chain, trim the sequence to V-start -> J-end (same coordinates
  # used by the tree builder) and count how many cells in the same clone share
  # an identical VDJ DNA sequence. This becomes Duplicate_DNA_H / _L in the
  # wide-format table and tells you how many cells made the exact same antibody.
  bcr <- bcr %>%
    dplyr::mutate(
      VDJ_DNA = dplyr::if_else(
        !is.na(v_sequence_start) & !is.na(j_sequence_end) &
          v_sequence_start > 0   & j_sequence_end > 0,
        substr(sequence, v_sequence_start, j_sequence_end),
        sequence
      )
    ) %>%
    dplyr::group_by(clone_id, locus, VDJ_DNA) %>%
    dplyr::mutate(
      Duplicate_DNA_H = dplyr::if_else(locus == "IGH",             dplyr::n(), NA_integer_),
      Duplicate_DNA_L = dplyr::if_else(locus %in% c("IGK", "IGL"), dplyr::n(), NA_integer_)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(-VDJ_DNA)

  bcr_annotated[[sid]] <- bcr
  message("  Done: ", nrow(bcr), " total contigs | ",
          length(unique(bcr$cell_id[bcr$locus == "IGH"])), " cells with heavy chain")
}
```

---

### 8.2a Save the annotated long-format BCR data

We save the annotated BCR data for each sample as a TSV. This is the
authoritative input for all BCR visualization downstream, and also feeds into
Step 9. If you need to re-run any plots without re-loading the Seurat object,
you can load these files directly and skip Steps 8.1--8.2.

```{r save-long-format}
for (sid in names(bcr_annotated)) {
  out_path <- file.path(out_dir, paste0(sid, "_S8_bcr_annotated.tsv"))
  data.table::fwrite(bcr_annotated[[sid]], file = out_path, sep = "\t")
  message("Saved: ", out_path)
}
```

---

### 8.2b Build and save the wide-format summary table

The long-format BCR data has two rows per cell (one for the heavy chain, one
for the light chain). For inspection and tabular analysis it is useful to have
a single-row-per-cell summary with both chains side by side. We build this here
and save it as an Excel file.

The table includes: cell barcode, clone ID, clone size, sample, cell type,
isotype, light chain type (kappa/lambda), VDJ duplicate counts for heavy and
light chains, somatic hypermutation counts and frequencies for both chains,
V/D/J gene calls, and the full nucleotide sequences.

```{r wide-format-table}
for (sid in names(bcr_annotated)) {

  bcr <- bcr_annotated[[sid]]

  heavy <- bcr %>%
    dplyr::filter(locus == "IGH") %>%
    dplyr::transmute(
      cell_id,
      clone_id,
      clone_count,
      sample_id,
      cell_type,
      isotype        = c_call,
      Duplicate_DNA_H,
      mu_count_H     = mu_count,
      mu_freq_H      = mu_freq,
      v_call_H       = v_call,
      d_call_H       = d_call,
      j_call_H       = j_call,
      sequence_H     = sequence
    )

  light <- bcr %>%
    dplyr::filter(locus %in% c("IGK", "IGL")) %>%
    dplyr::transmute(
      cell_id,
      light_chain = dplyr::case_when(
        locus == "IGK" ~ "Kappa",
        locus == "IGL" ~ "Lambda",
        TRUE           ~ locus
      ),
      Duplicate_DNA_L,
      mu_count_L = mu_count,
      mu_freq_L  = mu_freq,
      v_call_L   = v_call,
      j_call_L   = j_call,
      sequence_L = sequence
    )

  wide <- dplyr::left_join(heavy, light, by = "cell_id") %>%
    dplyr::arrange(dplyr::desc(clone_count))

  out_path <- file.path(out_dir, paste0(sid, "_S8_final_table.xlsx"))
  writexl::write_xlsx(wide, out_path)
  message(sid, ": wide-format table saved (", nrow(wide), " cells) -- ", out_path)
}
```

---

## 8.3 Donut plots

Donut plots are one of the most intuitive ways to visualize clonal structure in
BCR data. Each slice represents one clone, with arc length proportional to the
number of cells in that clone. Singletons (clones with only one cell) are
aggregated into a single white arc -- they correctly represent the proportion of
the repertoire that is non-expanded, but individual singleton clones carry no
meaningful information at this level of analysis.

Expanded clones are colored by their dominant heavy chain isotype using the
standard palette defined in `BCR_viz_functions.R`, giving an immediate visual
read of the isotype composition of the expanded repertoire.

Three levels of granularity are shown:

- **Combined** -- one donut for the whole sample showing all clones. Each slice
  represents one expanded clone colored by its heavy chain isotype; singletons
  are aggregated into a single white arc. The relative size of the white arc
  vs the colored arcs gives an immediate read of how clonal the repertoire is.
- **Per isotype** -- one donut per isotype, showing clonal structure within
  each isotype class. Each expanded clone gets a shade within that isotype's
  color family; singletons are white.
- **Per cluster** -- one donut per annotated cell type, showing the clonal
  landscape within each population. BCR cells with no Seurat match (NA cell
  type) are excluded here since they have no cluster assignment. In this
  dataset, PC clusters show the highest clonality. Naive B cells show the
  lowest clonality -- near-zero expansion is expected for this population. GC
  clusters can also show substantial expansion. MBC subsets vary.

---

### 8.3.1 Combined donut -- P1_LN

```{r donut-combined-ln, fig.height=6, fig.width=6}
donuts_ln <- plot_combined_donut(bcr_annotated[["P1_LN"]], sample_name = "P1_LN")
print(donuts_ln$all_clones)
```

---

### 8.3.2 Combined donut -- P1_PT

```{r donut-combined-pt, fig.height=6, fig.width=6}
donuts_pt <- plot_combined_donut(bcr_annotated[["P1_PT"]], sample_name = "P1_PT")
print(donuts_pt$all_clones)
```

---

### 8.3.3 Per-isotype donuts -- P1_LN

Each plot shows the clonal structure within one isotype. Expanded clones are
shown in shades of the isotype's color family (shades of green for IGHG1,
shades of blue for IGHM etc.), with darker shades indicating larger clones.
Singletons are white. Isotypes with fewer than 5 cells are skipped.

```{r donut-isotype-ln, fig.height=6, fig.width=6}
iso_donuts_ln <- plot_isotype_donuts(bcr_annotated[["P1_LN"]], sample_name = "P1_LN")

for (iso in names(iso_donuts_ln)) {
  print(iso_donuts_ln[[iso]]$all_clones)
}
```

---

### 8.3.4 Per-isotype donuts -- P1_PT

```{r donut-isotype-pt, fig.height=6, fig.width=6}
iso_donuts_pt <- plot_isotype_donuts(bcr_annotated[["P1_PT"]], sample_name = "P1_PT")

for (iso in names(iso_donuts_pt)) {
  print(iso_donuts_pt[[iso]]$all_clones)
}
```

---

### 8.3.5 Per-cluster donuts -- P1_LN

Clusters with fewer than 10 cells are skipped automatically. BCR cells that
did not match a Seurat barcode (NA cell type) are excluded from this section.

```{r donut-cluster-ln, fig.height=6, fig.width=6}
cluster_donuts_ln <- plot_cluster_donuts(
  bcr_annotated[["P1_LN"]],
  sample_name   = "P1_LN",
  cell_type_col = "cell_type"
)

for (cl in names(cluster_donuts_ln)) {
  print(cluster_donuts_ln[[cl]]$all_clones)
}
```

---

### 8.3.6 Per-cluster donuts -- P1_PT

```{r donut-cluster-pt, fig.height=6, fig.width=6}
cluster_donuts_pt <- plot_cluster_donuts(
  bcr_annotated[["P1_PT"]],
  sample_name   = "P1_PT",
  cell_type_col = "cell_type"
)

for (cl in names(cluster_donuts_pt)) {
  print(cluster_donuts_pt[[cl]]$all_clones)
}
```

---

## 8.4 Phylogenetic trees

Phylogenetic trees reconstruct the evolutionary history of a B cell clone --
how cells diversified from a common unmutated ancestor through rounds of
somatic hypermutation in the germinal center. Unlike generic phylogenetic tools,
dowser builds trees specifically designed for B cell clonal evolution. The key
difference is that the unmutated germline sequence (reconstructed in Part 1 by
`createGermlines()`) is used as the known root of the tree. This means the root
is not inferred from the data but is fixed to the true biological ancestor,
giving the tree a meaningful directionality: branch length represents mutations
accumulated from the germline, and the root-to-tip direction represents time
and affinity maturation.

Each tip represents one unique sequence (or a group of cells sharing an
identical sequence, collapsed to reduce redundancy). Tip size reflects how many
cells share that sequence. Tips are colored by isotype, showing how isotype
switching occurred within the clone. Cell type labels are shown as colored
rectangles next to tips where the cell type is known and at least two cells
share that sequence -- singleton tips are shown as points only to avoid
cluttering small trees.

We build trees for the top 5 expanded clones per sample by default. You can
change `top_n` or supply specific clone IDs via `clone_ids` to inspect
particular clones of interest.

---

### 8.4a Cluster color palette

The tree tip labels use the same cell type color palette as the UMAP in Part 2.

```{r cluster-colors}
cluster_colors <- c(
  "MBC_Basal"    = "#F8766D",
  "MBC_Atypical" = "#D89000",
  "Naive"        = "#A3A500",
  "PC"           = "#39B600",
  "GC"           = "#00BF7D",
  "GC_Cycling"   = "#00BFC4",
  "MBC_CR2pos"   = "#00B0F6",
  "PC_2"         = "#9590FF",
  "Plasmablast"  = "#E76BF3",
  "Unsure"       = "#FF62BC"
)
```

---

### 8.4.1 Trees -- P1_LN

```{r trees-ln, fig.height=8, fig.width=10}
# If IQ-TREE 2 is not on your system PATH, supply the full path:
#   Windows : exec = "C:/tools/iqtree2/bin/iqtree2.exe"
#   macOS   : exec = "/usr/local/bin/iqtree2"

tree_plots_ln <- build_bcr_trees(
  bcr_data       = bcr_annotated[["P1_LN"]],
  cluster_colors = cluster_colors,
  top_n          = 5,
  cell_type_col  = "cell_type",
  exec           = "iqtree2"
)

for (cid in names(tree_plots_ln)) {
  if (!is.null(tree_plots_ln[[cid]])) {
    print(tree_plots_ln[[cid]])
  }
}
```

---

### 8.4.2 Trees -- P1_PT

```{r trees-pt, fig.height=8, fig.width=10}
tree_plots_pt <- build_bcr_trees(
  bcr_data       = bcr_annotated[["P1_PT"]],
  cluster_colors = cluster_colors,
  top_n          = 5,
  cell_type_col  = "cell_type",
  exec           = "iqtree2"
)

for (cid in names(tree_plots_pt)) {
  if (!is.null(tree_plots_pt[[cid]])) {
    print(tree_plots_pt[[cid]])
  }
}
```

---

### 8.4.3 Building a tree for a specific clone

To build a tree for a particular clone of interest rather than the top N by
size, supply the clone ID directly via the `clone_ids` argument. Clone IDs
follow the format `<sample_id>_<random_code>_<clone_size>_<isotypes>` and can
be found in the wide-format Excel table saved in Section 8.2b.

```{r trees-specific, eval=FALSE}
# Replace with your clone ID of interest
target_clone <- "P1_LN_4aFs_6_G2G4"

tree_specific <- build_bcr_trees(
  bcr_data       = bcr_annotated[["P1_LN"]],
  cluster_colors = cluster_colors,
  clone_ids      = target_clone,
  cell_type_col  = "cell_type",
  exec           = "iqtree2"
)

print(tree_specific[[target_clone]])
```

---

### 8.4.4 Save tree plots to PDF

```{r save-trees}
tree_dir <- file.path(plot_dir, "Trees")
dir.create(tree_dir, recursive = TRUE, showWarnings = FALSE)

for (sid in c("P1_LN", "P1_PT")) {
  plots <- if (sid == "P1_LN") tree_plots_ln else tree_plots_pt

  for (cid in names(plots)) {
    if (!is.null(plots[[cid]])) {
      out_file <- file.path(tree_dir, paste0(cid, "_tree.pdf"))
      ggplot2::ggsave(out_file, plots[[cid]], width = 10, height = 8)
      message("Saved: ", out_file)
    }
  }
}

message("All tree plots saved.")
```

---

# Step 9: Cross-Sample Clonal Analysis with scRepertoire

The analyses in Step 8 are per-sample: clones were defined independently for
each sample in Part 1, so a clone in P1_LN and a clone in P1_PT with identical
V(D)J sequences are treated as separate lineages. Step 9 asks a different
question: are any clones shared between the LN and PT samples from the same
patient? To answer this we use scRepertoire, which applies a cross-sample
clonotype definition (CTstrict) based on shared V gene usage and greater than
85% CDR3 nucleotide similarity. This is intentionally distinct from the
Immcantation clone IDs used in Step 8. The two definitions serve different
purposes and should not be mixed.

**A note on clonotype definitions:** scRepertoire's CTstrict and Immcantation's
`hierarchicalClones()` are not equivalent. Immcantation uses a mixture-model
distance threshold on full heavy-chain sequences with light-chain splitting;
scRepertoire uses a fixed 85% CDR3 similarity cutoff. CTstrict is specifically
designed for cross-sample comparison where you do not want to re-run Immcantation
on the combined data. Do not replace the Immcantation clone IDs with CTstrict
for within-sample analyses -- they serve different purposes.

---

## 9.1 Prepare AIRR data for scRepertoire

scRepertoire's `combineBCR()` expects a named list of data frames, one per
sample, with specific column names. Our Immcantation AIRR data uses slightly
different names (e.g. `cell_id` instead of `barcode`, `locus` instead of
`chain`, `v_call` instead of `v_gene`). The function below maps between them.

Note that `cdr3` (amino acid) and `cdr3_nt` (nucleotide) are both mapped from
the `junction` and `junction_aa` columns. scRepertoire uses `cdr3_nt` for the
CTstrict distance calculation.

```{r prepare-screp}
prepare_for_scRepertoire <- function(df) {

  # barcode and sample identity
  if (!"barcode"   %in% names(df) && "cell_id"   %in% names(df)) df$barcode   <- df$cell_id
  if (!"sample"    %in% names(df) && "sample_id" %in% names(df)) df$sample    <- df$sample_id

  # chain locus
  if (!"chain"     %in% names(df) && "locus"     %in% names(df)) df$chain     <- df$locus

  # gene calls
  if (!"v_gene"    %in% names(df) && "v_call"    %in% names(df)) df$v_gene    <- df$v_call
  if (!"d_gene"    %in% names(df) && "d_call"    %in% names(df)) df$d_gene    <- df$d_call
  if (!"j_gene"    %in% names(df) && "j_call"    %in% names(df)) df$j_gene    <- df$j_call
  if (!"c_gene"    %in% names(df) && "c_call"    %in% names(df)) df$c_gene    <- df$c_call

  # CDR3 sequences -- scRepertoire uses cdr3 (aa) and cdr3_nt for CTstrict
  if (!"cdr3"      %in% names(df) && "junction_aa" %in% names(df)) df$cdr3    <- df$junction_aa
  if (!"cdr3_nt"   %in% names(df) && "junction"    %in% names(df)) df$cdr3_nt <- df$junction
  if (!"cdr3_aa"   %in% names(df) && "junction_aa" %in% names(df)) df$cdr3_aa <- df$junction_aa

  # UMI count as a reads proxy (optional but avoids NAs in scRepertoire internals)
  if (!"reads"     %in% names(df) && "umi_count"   %in% names(df)) df$reads   <- df$umi_count
  if (!"reads"     %in% names(df))                                  df$reads   <- 1L

  return(df)
}
```

Now load both samples, apply the mapping, and run `combineBCR()`.
`removeMulti = TRUE` drops cells with more than one heavy chain (these were
already filtered in Part 1, so this is a safety net only).

```{r combine-bcr}
# Load the annotated BCR TSVs saved in Section 8.2a
annotated_tsv <- list(
  P1_LN = file.path(out_dir, "P1_LN_S8_bcr_annotated.tsv"),
  P1_PT = file.path(out_dir, "P1_PT_S8_bcr_annotated.tsv")
)

airr_list <- lapply(annotated_tsv, function(path) {
  if (!file.exists(path)) stop("File not found: ", path)
  data.table::fread(path, sep = "\t", data.table = FALSE)
})

airr_list <- lapply(airr_list, prepare_for_scRepertoire)

# Quick check: confirm required columns are present
required_cols <- c("barcode", "chain", "v_gene", "j_gene", "cdr3_nt")
for (sid in names(airr_list)) {
  missing <- setdiff(required_cols, colnames(airr_list[[sid]]))
  if (length(missing) > 0)
    stop(sid, ": missing required columns after mapping: ",
         paste(missing, collapse = ", "))
}

message("Running combineBCR() across ", length(airr_list), " samples...")

# Note on barcode format: combineBCR() prepends the sample name to each
# barcode, producing e.g. "P1_LN_ACTGCTCA...-1_P1_LN". This double
# suffix/prefix is expected -- our cell_ids already carry the sample suffix
# from Part 1, and combineBCR() adds its own prefix on top. The ct_lookup
# step in 9.3.1 strips the leading prefix to recover the original cell_id
# format used throughout Parts 1-8.

combined_BCR <- combineBCR(
  airr_list,
  samples     = names(airr_list),
  removeNA    = FALSE,
  removeMulti = TRUE
)

# Confirm output
message("combineBCR() complete.")
message("Cells per sample:")
print(sapply(combined_BCR, nrow))
message("Columns in combined_BCR output:")
print(colnames(combined_BCR[[1]]))
message("Example CTstrict values (P1_LN):")
print(head(combined_BCR[["P1_LN"]]$CTstrict, 10))

# Save
saveRDS(combined_BCR, file.path(rds_dir, "combined_BCR_S9.rds"))
message("Saved: combined_BCR_S9.rds")
```

---

## 9.2 Repertoire-level characterization

Before looking at shared clones, we characterize each sample's repertoire
structure independently.

---

### 9.2.1 Clonal homeostasis

`clonalHomeostasis()` shows what fraction of each sample's repertoire is
occupied by clones in different size classes. Each bar represents one sample;
the stacked segments represent the proportion of cells belonging to rare,
small, medium, large, and hyperexpanded clones. This gives an immediate read
of how oligoclonal each repertoire is.

For B cell data the default proportional bins (Rare < 0.0001, Small < 0.001,
Medium < 0.01, Large < 0.1, Hyperexpanded < 1) are appropriate.

In this patient, PT shows substantially more clonal expansion than LN, with a
notable fraction of cells in the Medium and Large categories. LN is almost
entirely composed of Small and Rare clones.

```{r homeostasis, fig.height=5, fig.width=6}
p_homeostasis <- clonalHomeostasis(
  combined_BCR,
  cloneCall = "strict"
)

p_homeostasis

ggsave(file.path(plot_dir, "clonal_homeostasis.pdf"),
       p_homeostasis, width = 6, height = 5)
message("Saved: clonal_homeostasis.pdf")
```

---

## 9.3 Quantify shared clones

Here we break down the repertoire overlap into specific numbers: how many
CTstrict clones are shared between LN and PT, and how many cells belong to
those shared clones. We also build a master table that joins CTstrict onto the
per-cell BCR data from Step 8.

**A note on method and reproducibility:** This tutorial uses pure CTstrict for
both identifying shared clones and counting cells within them. If you compare
these numbers against a paper that used a hybrid approach -- using CTstrict to
identify shared clones but Immcantation `clone_id` to count cells -- you will
get different totals. Immcantation groups cells more aggressively than CTstrict,
so the hybrid approach typically counts more cells per shared clone. Neither is
wrong; they answer slightly different questions. The pure CTstrict approach used
here is more reproducible because it depends only on scRepertoire.

**A note on scRepertoire version:** This tutorial was written and tested against
scRepertoire v1.12.0. The CTstrict string format used to identify shared clones
changed between major versions, so the same input data can produce slightly
different shared clone counts depending on which version is installed. With
v1.12.0, the shared count for P1 in Section 9.3.3 below is 151. On other
versions you may see a different number for the same set of cells and the same
underlying Immcantation clones -- the cells matched are identical, only the
string used to compare them is formatted differently. To exactly reproduce the
counts shown here, pin scRepertoire to v1.12.0. Otherwise small numeric
differences are expected and do not reflect a bug.

---

### 9.3.1 Build the CTstrict lookup

`combineBCR()` prepends the sample name to each barcode (e.g.
`P1_LN_ACGT...-1_P1_LN`). The BCR data from Step 8 uses `cell_id` in the
format `ACGT...-1_P1_LN`. We strip the leading `<sample>_` prefix from the
combined_BCR barcodes so they match the Step 8 cell_id format.

```{r ct-lookup}
ct_lookup <- do.call(rbind, lapply(names(combined_BCR), function(sid) {
  df <- combined_BCR[[sid]]
  data.frame(
    sample_id = df$sample,
    # Remove the leading "<sample>_" prefix that combineBCR adds
    cell_id   = sub(paste0("^", sid, "_"), "", df$barcode),
    CTstrict  = df$CTstrict,
    CTaa      = df$CTaa,
    CTnt      = df$CTnt,
    CTgene    = df$CTgene,
    stringsAsFactors = FALSE
  )
}))

message("CTstrict lookup: ", nrow(ct_lookup), " rows")
message("Example cell_id values after stripping:")
print(head(ct_lookup$cell_id, 6))
```

---

### 9.3.2 Join CTstrict onto the master BCR table

Load both annotated BCR TSVs from Step 8, combine them, and join the CTstrict
column. This master table is the single input for the shared clone dot plot
in Step 9.6.

```{r master-table}
bcr_all <- do.call(rbind, lapply(names(annotated_tsv), function(sid) {
  data.table::fread(annotated_tsv[[sid]], sep = "\t", data.table = FALSE)
}))

message("Master BCR table: ", nrow(bcr_all), " rows | ",
        length(unique(bcr_all$cell_id[bcr_all$locus == "IGH"])),
        " cells with heavy chain")

# Left-join CTstrict by sample_id + cell_id
bcr_all <- dplyr::left_join(
  bcr_all,
  ct_lookup,
  by = c("sample_id", "cell_id")
)

# Check match rate
n_igh     <- sum(bcr_all$locus == "IGH")
n_matched <- sum(!is.na(bcr_all$CTstrict[bcr_all$locus == "IGH"]))
message("CTstrict matched: ", n_matched, " / ", n_igh,
        " heavy chain cells (",
        round(100 * n_matched / n_igh, 1), "%)")

# If the match rate is below 50%, check that the barcode format is consistent
# between Part 1 and Part 2. The cell_id in Step 8 should look like
# "ACGT...-1_P1_LN" and the stripped combined_BCR barcode should match exactly.
if (n_matched / n_igh < 0.5) {
  warning("Less than 50% of heavy chain cells matched a CTstrict. ",
          "Check barcode format consistency between Part 1 and Part 2.")
}

# Save
data.table::fwrite(bcr_all,
                   file.path(out_dir, "All_S9_bcr_with_CTstrict.tsv"),
                   sep = "\t")
message("Saved: All_S9_bcr_with_CTstrict.tsv")
```

---

### 9.3.3 Count shared clones and cells

```{r shared-counts}
# Tissue label (LN or PT) derived from sample_id
bcr_all <- bcr_all %>%
  dplyr::mutate(
    tissue = dplyr::case_when(
      grepl("_LN$|^LN", sample_id) ~ "LN",
      grepl("_PT$|^PT", sample_id) ~ "PT",
      TRUE ~ "UNK"
    )
  )

# Restrict to heavy chain rows with a valid CTstrict
bcr_igh <- bcr_all %>%
  dplyr::filter(locus == "IGH", !is.na(CTstrict), CTstrict != "")

# Which CTstrict clones appear in both LN and PT?
clones_ln <- unique(bcr_igh$CTstrict[bcr_igh$tissue == "LN"])
clones_pt <- unique(bcr_igh$CTstrict[bcr_igh$tissue == "PT"])
shared_cts <- intersect(clones_ln, clones_pt)

message("CTstrict clones in LN only:   ", length(setdiff(clones_ln, clones_pt)))
message("CTstrict clones in PT only:   ", length(setdiff(clones_pt, clones_ln)))
message("CTstrict clones shared LN+PT: ", length(shared_cts))

# Cells per tissue belonging to shared clones
# (raw counts only -- no percentages; denominator choice is non-trivial)
cells_ln_shared <- sum(bcr_igh$CTstrict[bcr_igh$tissue == "LN"] %in% shared_cts)
cells_pt_shared <- sum(bcr_igh$CTstrict[bcr_igh$tissue == "PT"] %in% shared_cts)

message("LN cells in shared clones:    ", cells_ln_shared)
message("PT cells in shared clones:    ", cells_pt_shared)
message("Total shared cells:           ", cells_ln_shared + cells_pt_shared)

# Summary table of shared clones with cell counts per tissue
shared_clone_tbl <- bcr_igh %>%
  dplyr::filter(CTstrict %in% shared_cts) %>%
  dplyr::count(CTstrict, tissue, name = "n_cells") %>%
  tidyr::pivot_wider(
    names_from  = tissue,
    values_from = n_cells,
    values_fill = 0L
  ) %>%
  dplyr::mutate(total_cells = LN + PT) %>%
  dplyr::arrange(dplyr::desc(total_cells))

message("\nShared clone table (top 10):")
print(knitr::kable(head(shared_clone_tbl, 10), caption = "Top 10 shared clones by total cell count"))
```

---

### 9.3.4 Visualize shared vs. exclusive cell counts

With only two samples, summary statistics like the Morisita overlap index
reduce to a single number and are not informative as a standalone plot.
Instead, we show a stacked bar chart of cell counts broken down by tissue
and sharing status: how many IGH cells in each sample belong to
tissue-exclusive clones vs. clones shared with the other tissue.

```{r shared-bar, fig.height=4, fig.width=5}
shared_bar_df <- bcr_igh %>%
  dplyr::mutate(
    status = dplyr::if_else(CTstrict %in% shared_cts, "Shared", "Exclusive")
  ) %>%
  dplyr::count(tissue, status, name = "n_cells") %>%
  dplyr::mutate(
    tissue = factor(tissue, levels = c("LN", "PT")),
    status = factor(status, levels = c("Shared", "Exclusive"))
  )

p_shared_bar <- ggplot(shared_bar_df,
                       aes(x = tissue, y = n_cells, fill = status)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = n_cells),
            position = position_stack(vjust = 0.5),
            color = "white", size = 4, fontface = "bold") +
  scale_fill_manual(
    values = c(Shared = "#2166AC", Exclusive = "#B2B2B2"),
    name   = "Clone status"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank()) +
  labs(
    x     = "Tissue",
    y     = "Number of cells (IGH)",
    title = "Cells in shared vs. exclusive clones -- P1"
  )

p_shared_bar

ggsave(file.path(plot_dir, "shared_cells_bar_P1.pdf"),
       p_shared_bar, width = 5, height = 4)
message("Saved: shared_cells_bar_P1.pdf")
```

---

## 9.4 Attach clonal data to the Seurat object

`combineExpression()` adds the scRepertoire clonotype metadata to the Seurat
object's `@meta.data` slot. After this step, every cell in the UMAP has a
CTstrict label, a cloneSize category, and a clonal frequency, enabling
visualization of clonal structure directly on the dimensional reduction.

The barcode format in the Seurat object is `ACGT...-1_P1_LN` (barcode + sample
suffix, added during integration in Part 2). The format in combined_BCR after
`combineBCR()` is `P1_LN_ACGT...-1_P1_LN` (sample prefix + barcode + original
suffix). We harmonize these before calling `combineExpression()` by stripping
the leading `<sample>_` prefix from combined_BCR barcodes.

```{r attach-seurat}
seu <- readRDS(rds_in)
message("Seurat object loaded: ", ncol(seu), " cells")

# Harmonize barcodes: strip leading "<sample>_" prefix from combined_BCR
combined_BCR_harmonized <- lapply(combined_BCR, function(df) {
  df$barcode <- sub(paste0("^", df$sample, "_"), "", df$barcode)
  df
})

# Overlap check before proceeding
airr_barcodes <- unique(unlist(lapply(combined_BCR_harmonized,
                                       function(x) x$barcode)))
n_overlap <- sum(colnames(seu) %in% airr_barcodes)
message("Barcode overlap (Seurat intersect AIRR): ", n_overlap, " / ", ncol(seu))

if (n_overlap == 0) {
  stop(
    "Zero overlap after barcode harmonization.\n",
    "Check that cell_id format in Step 8 TSVs matches Seurat colnames.\n",
    "Example Seurat names: ", paste(head(colnames(seu), 3), collapse = ", "), "\n",
    "Example AIRR barcodes: ", paste(head(airr_barcodes, 3), collapse = ", ")
  )
}

# Add a 'sample' column to Seurat metadata for combineExpression grouping.
# Seurat cell names look like "ACGT...-1_P1_LN". We need the sample label to
# match the names of combined_BCR_harmonized (e.g. "P1_LN", "P1_PT").
# We extract the suffix after the last "-1_" by matching known sample IDs.
sample_ids <- names(combined_BCR_harmonized)
seu$sample <- NA_character_
for (sid in sample_ids) {
  seu$sample[grepl(paste0("_", sid, "$"), colnames(seu))] <- sid
}
message("Seurat sample labels:")
print(table(seu$sample, useNA = "ifany"))

# Run combineExpression
# proportion = TRUE is required so Frequency is stored as a proportion.
# The cloneType bins (Rare/Small/Medium/Large/Hyperexpanded) use proportion
# thresholds (0.0001 / 0.001 / 0.01 / 0.1 / 1). With proportion = FALSE,
# raw counts are stored and every clone exceeds the 0.1 threshold, collapsing
# all cells into "Hyperexpanded".
# group.by is intentionally omitted: in scRepertoire v2, combining group.by
# with proportion = TRUE causes frequencies to be calculated incorrectly,
# also resulting in all cells landing in "Hyperexpanded".
seu_scRep <- combineExpression(
  combined_BCR_harmonized,
  sc         = seu,
  cloneCall  = "strict",
  proportion = TRUE,
  filterNA   = FALSE
)

message("combineExpression() complete.")
message("New metadata columns added:")
new_cols <- setdiff(colnames(seu_scRep@meta.data), colnames(seu@meta.data))
print(new_cols)

# Confirm CTstrict is present
if (!"CTstrict" %in% colnames(seu_scRep@meta.data)) {
  warning("CTstrict column not found in Seurat metadata. ",
          "Check that cloneCall = 'strict' matches the combined_BCR columns.")
} else {
  n_with_ct <- sum(!is.na(seu_scRep$CTstrict))
  message("Cells with CTstrict assigned: ", n_with_ct, " / ", ncol(seu_scRep))
}

# Save
saveRDS(seu_scRep, file.path(rds_dir, "integrated_S9_scRep.rds"))
message("Saved: integrated_S9_scRep.rds")
```

---

## 9.5 Single-cell visualizations

---

### 9.5.1 Clonal network on UMAP

`clonalNetwork()` draws arrows between clusters that share CTstrict clones,
with arrow thickness proportional to the number of cells involved. The overall
UMAP layout is preserved so you can see which cell types are clonally connected.

We first plot all connections across the whole object, then filter to a single
cluster of interest using `filter.identity`. This is useful for asking which
other clusters share clones with, say, the GC or plasma cell population.

```{r clonal-network-all, fig.height=7, fig.width=9}
set.seed(42)  # ensures reproducibility of clonalNetwork() force-directed layout (igraph is stochastic)
# Full network -- all clusters
p_net_all <- clonalNetwork(
  seu_scRep,
  reduction       = "umap",
  identity        = "cell_type",
  filter.clones   = NULL,
  filter.identity = NULL,
  cloneCall       = "strict"
)

p_net_all

ggsave(file.path(plot_dir, "clonal_network_all.pdf"),
       p_net_all, width = 9, height = 7)
message("Saved: clonal_network_all.pdf")
```

Now filter to a single cluster of interest. Replace `"PC"` below with
whichever cell type you want to examine. The plot will show only the
connections involving that cluster.

```{r clonal-network-filtered, fig.height=7, fig.width=9}
set.seed(42)  # ensures reproducibility of clonalNetwork() force-directed layout (igraph is stochastic)
# Replace "PC" with your cluster of interest
cluster_of_interest <- "PC"

p_net_filtered <- clonalNetwork(
  seu_scRep,
  reduction       = "umap",
  identity        = "cell_type",
  filter.identity = cluster_of_interest,
  filter.clones   = NULL,
  cloneCall       = "strict"
)

p_net_filtered

ggsave(file.path(plot_dir,
                 paste0("clonal_network_", cluster_of_interest, ".pdf")),
       p_net_filtered, width = 9, height = 7)
message("Saved: clonal_network_", cluster_of_interest, ".pdf")
```

---

### 9.5.2 Clonal occupancy by cluster

`occupiedscRepertoire()` shows how many cells in each cluster belong to each
cloneSize category (Single, Small, Medium, Large, Hyperexpanded), as defined
by `combineExpression()`. This is a useful complement to the donut plots in
Step 8: instead of showing clone identity, it shows the expansion profile of
each annotated cell type.

Note: `clonalOccupy()` was renamed to `occupiedscRepertoire()` in scRepertoire
v2.0. If you get a "could not find function" error for one, use the other.

```{r clonal-occupy, fig.height=5, fig.width=9}
# Set cell_type factor levels to control x-axis order
cluster_order <- c(
  "Naive", "MBC_Basal", "MBC_CR2pos", "MBC_Atypical",
  "GC", "GC_Cycling", "PC", "PC_2", "Plasmablast", "Unsure"
)

seu_scRep$cell_type <- factor(seu_scRep$cell_type, levels = cluster_order)

p_occupy <- occupiedscRepertoire(
  seu_scRep,
  x.axis     = "cell_type",
  proportion = FALSE,
  label      = TRUE
)

p_occupy +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(plot_dir, "clonal_occupy_counts.pdf"),
       p_occupy + theme(axis.text.x = element_text(angle = 45, hjust = 1)),
       width = 9, height = 5)

# Proportion version -- useful when cluster sizes differ greatly
p_occupy_prop <- occupiedscRepertoire(
  seu_scRep,
  x.axis     = "cell_type",
  proportion = TRUE,
  label      = FALSE
)

p_occupy_prop +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(plot_dir, "clonal_occupy_proportion.pdf"),
       p_occupy_prop + theme(axis.text.x = element_text(angle = 45, hjust = 1)),
       width = 9, height = 5)

message("Saved: clonal_occupy_counts.pdf and clonal_occupy_proportion.pdf")
```

---

## 9.6 Shared clone dot plot

This plot shows which cell type clusters each shared LN/PT clone occupies.
Each row is one shared CTstrict clone, ordered by total cell count (largest
at top). Each column is a cell type cluster. Each dot encodes three things:

- **Color** -- dominant heavy chain isotype for that clone x cluster combination
- **Shape** -- tissue of origin: circle = LN only, square = PT only,
  triangle = cells from both LN and PT found in that cluster
- **Size** -- number of cells in that clone x cluster combination

Only clones with at least one cell in both LN and PT (after restricting to
cells with a valid `cell_type` annotation) are shown.

---

### 9.6a Color palette and helper functions

```{r shared-plot-setup}
# Isotype colors -- consistent with the donut palette in Step 8
isotype_colors <- c(
  IGHD  = "#E41A1C",
  IGHM  = "#377EB8",
  IGHG1 = "#4DAF4A",
  IGHG2 = "#984EA3",
  IGHG3 = "#FF7F00",
  IGHG4 = "#FFFF33",
  IGHA1 = "#A65628",
  IGHA2 = "#F781BF"
)

isotype_priority <- c("IGHG1", "IGHG2", "IGHG3", "IGHG4",
                      "IGHA1", "IGHA2", "IGHD",  "IGHM")

# For each clone x cluster, pick the most common isotype.
# Ties are broken by the priority order above (IgG first, then IgA, IgD, IgM).
pick_dominant_isotype <- function(x) {
  x <- x[!is.na(x) & x != "IGHE"]
  if (length(x) == 0) return(NA_character_)
  tab  <- table(x)
  mx   <- max(tab)
  tied <- names(tab)[tab == mx]
  if (length(tied) == 1) return(tied)
  tied <- intersect(isotype_priority, tied)
  if (length(tied) > 0) return(tied[1])
  sort(names(tab)[tab == mx])[1]
}
```

---

### 9.6b Build the plot data

```{r shared-plot-data}
# Use only heavy chain rows that have a valid CTstrict, cell_type, and isotype
bcr_plot <- bcr_all %>%
  dplyr::filter(
    locus     == "IGH",
    !is.na(CTstrict),  CTstrict  != "",
    !is.na(cell_type), cell_type != "",
    !is.na(c_call),    c_call    != "",
    c_call    != "IGHE"
  ) %>%
  dplyr::mutate(
    tissue = dplyr::case_when(
      sample_id == "P1_LN" ~ "LN",
      sample_id == "P1_PT" ~ "PT",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(tissue))

# Identify clones present in both LN and PT
shared_cts_p1 <- bcr_plot %>%
  dplyr::group_by(CTstrict) %>%
  dplyr::summarise(
    has_LN = any(tissue == "LN"),
    has_PT = any(tissue == "PT"),
    .groups = "drop"
  ) %>%
  dplyr::filter(has_LN & has_PT) %>%
  dplyr::pull(CTstrict)

message("Shared CTstrict clones with cell type annotation: ",
        length(shared_cts_p1))

# Collapse to one row per (CTstrict x cell_type)
plot_df <- bcr_plot %>%
  dplyr::filter(CTstrict %in% shared_cts_p1) %>%
  dplyr::group_by(CTstrict, cell_type) %>%
  dplyr::summarise(
    Count       = dplyr::n(),
    has_LN_here = any(tissue == "LN"),
    has_PT_here = any(tissue == "PT"),
    ccall_dom   = pick_dominant_isotype(c_call),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    Shape = dplyr::case_when(
      has_LN_here & has_PT_here ~ "Both",
      has_PT_here               ~ "PT",
      has_LN_here               ~ "LN"
    ),
    ccall_dom  = factor(ccall_dom, levels = isotype_priority),
    cell_type  = factor(cell_type, levels = cluster_order),
    Count_plot = pmin(pmax(Count, 1L), max(Count))
  ) %>%
  dplyr::filter(!is.na(Shape), !is.na(ccall_dom))

# Order clones by total cell count -- largest clone at the top of the y axis
clone_order_tbl <- plot_df %>%
  dplyr::group_by(CTstrict) %>%
  dplyr::summarise(total_cells = sum(Count), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(total_cells))

plot_df <- plot_df %>%
  dplyr::mutate(CloneID = factor(CTstrict, levels = clone_order_tbl$CTstrict))

message("Plot data: ", dplyr::n_distinct(plot_df$CTstrict),
        " clones x ", dplyr::n_distinct(plot_df$cell_type), " clusters")
```

---

### 9.6c Build and save the plot

```{r shared-plot-build, fig.height=8, fig.width=10}
# Horizontal dotted guide lines -- one per clone row
y_lines <- seq_along(levels(plot_df$CloneID))

# Dynamic size scale: upper limit is the actual max cell count in the data
max_count <- max(plot_df$Count)
mid_count <- round(max_count / 2)

p_shared <- ggplot(plot_df,
                   aes(x     = cell_type,
                       y     = CloneID,
                       color = ccall_dom,
                       shape = Shape,
                       size  = Count_plot)) +
    geom_hline(yintercept = y_lines,
               linetype = "dotted", color = "grey70", linewidth = 0.4) +
    geom_point(stroke = 0.8) +
    scale_x_discrete(limits = cluster_order, drop = FALSE) +
    scale_color_manual(values = isotype_colors, name = "Isotype") +
    scale_shape_manual(
      values = c(PT = 15, LN = 16, Both = 17),
      breaks = c("PT", "LN", "Both"),
      labels = c(PT = "PT", LN = "LN", Both = "LN+PT"),
      name   = "Origin"
    ) +
    scale_size_continuous(
      limits = c(1, max_count),
      range  = c(3, 8),
      breaks = c(1, mid_count, max_count),
      labels = c("1", as.character(mid_count), as.character(max_count)),
      name   = "Cells / clone / cluster"
    ) +
    theme_minimal() +
    theme(
      legend.position    = "bottom",
      axis.text.x        = element_text(angle = 45, hjust = 1, size = 11),
      axis.text.y        = element_blank(),
      axis.ticks.y       = element_blank(),
      panel.grid.major.y = element_blank()
    ) +
    labs(
      x     = "Cell type cluster",
      y     = NULL,
      title = "Shared LN/PT clones -- P1"
    ) +
    guides(
      color = guide_legend(override.aes = list(size = 4)),
      shape = guide_legend(override.aes = list(size = 4)),
      size  = guide_legend(override.aes = list(shape = 16))
    )

p_shared

ggsave(file.path(plot_dir, "shared_clones_dotplot_P1.pdf"),
       p_shared, width = 10, height = 8)
message("Saved: shared_clones_dotplot_P1.pdf")
```

---

# Going further -- other scRepertoire analyses

Part 3 covers the analyses performed in the accompanying paper. scRepertoire
offers several additional analyses that may be useful depending on your dataset.
These are not run here -- with two samples many are underpowered -- but are
straightforward to add.

**Clonal diversity** -- compare Shannon entropy, inverse Simpson, Gini
coefficient, and other indices across samples. Most informative when comparing
many samples or conditions:

```r
clonalDiversity(combined_BCR, cloneCall = "strict", metric = "shannon")
clonalDiversity(combined_BCR, cloneCall = "strict", metric = "inv.simpson")
```

**V gene usage** -- visualize heavy chain V gene usage as a bar chart or
heatmap, or compare V-J pairings. Useful for identifying biased gene usage in
expanded clones:

```r
vizGenes(combined_BCR, x.axis = "IGHV", plot = "barplot",
         summary.fun = "proportion")
percentVJ(combined_BCR, chain = "IGH", summary.fun = "percent")
```

**Clonal proportion** -- rank-based view of how much repertoire space is
occupied by the top 10, top 100, top 1000 clones:

```r
clonalProportion(combined_BCR, cloneCall = "strict")
```

**Highlight specific clones on UMAP** -- pin a clone of interest and see
where its cells sit across the transcriptomic landscape:

```r
seu_scRep <- highlightClones(seu_scRep, cloneCall = "strict",
                              sequence = c("your_CTstrict_id"))
DimPlot(seu_scRep, group.by = "highlight")
```

**Alluvial plot** -- trace how clones flow across multiple categorical
variables (e.g. tissue, cell type, patient). Requires at least three
well-populated categories to be interpretable:

```r
alluvialClones(seu_scRep, clone.call = "strict",
               y.axes = c("sample", "cell_type"))
```

**Chord diagram** -- visualize the number of shared clones between clusters as
a circular diagram:

```r
library(circlize)
vizCirclize(seu_scRep, group.by = "cell_type")
```

For complete documentation and examples see the scRepertoire vignettes at
https://www.borch.dev/uploads/screpertoire/.

---

# Summary

| Output file | Contents |
|-------------|----------|
| `P1_LN_S8_bcr_annotated.tsv` | Long-format BCR data with cell_type joined, P1_LN |
| `P1_PT_S8_bcr_annotated.tsv` | Long-format BCR data with cell_type joined, P1_PT |
| `P1_LN_S8_final_table.xlsx` | Wide-format one-row-per-cell summary, P1_LN |
| `P1_PT_S8_final_table.xlsx` | Wide-format one-row-per-cell summary, P1_PT |
| `combined_BCR_S9.rds` | combineBCR() output -- named list of scRepertoire-ready data frames |
| `All_S9_bcr_with_CTstrict.tsv` | Master BCR table (both samples) with CTstrict joined |
| `integrated_S9_scRep.rds` | Seurat object with scRepertoire metadata (CTstrict, cloneSize, etc.) |
| `Plots/Part3/Trees/` | PDF tree plots for top 5 clones per sample |
| `Plots/Part3/clonal_homeostasis.pdf` | Clonal homeostasis stacked bar per sample |
| `Plots/Part3/shared_cells_bar_P1.pdf` | Shared vs. exclusive cell counts per tissue |
| `Plots/Part3/clonal_network_all.pdf` | clonalNetwork UMAP -- all clusters |
| `Plots/Part3/clonal_network_PC.pdf` | clonalNetwork UMAP -- PC cluster |
| `Plots/Part3/clonal_occupy_counts.pdf` | occupiedscRepertoire -- absolute counts |
| `Plots/Part3/clonal_occupy_proportion.pdf` | occupiedscRepertoire -- proportions |
| `Plots/Part3/shared_clones_dotplot_P1.pdf` | Shared clone dot plot -- Patient 1 |

---

*Part 3 complete. All BCR analysis outputs (per-sample annotation, clonal trees,
cross-sample sharing) and GEX outputs (integration, cell type annotation, UMAP)
are now in a single annotated Seurat object ready for downstream statistical
analysis and figure preparation.*
